High-Fidelity Neural Speech Reconstruction through an Efficient Acoustic-Linguistic Dual-Pathway Framework

Version 1

Abstract

Reconstructing speech from neural recordings is crucial for understanding speech coding and developing brain-computer interfaces (BCIs). However, existing methods trade off acoustic richness (pitch, prosody) for linguistic intelligibility (words, phonemes). To overcome this limitation, we propose a dual-path framework to concurrently decode acoustic and linguistic representations. The acoustic pathway uses a long-short term memory (LSTM) decoder and a high-fidelity generative adversarial network (HiFi-GAN) to reconstruct spectrotemporal features. The linguistic pathway employs a transformer adaptor and text-to-speech (TTS) generator for word tokens. These two pathways merge via voice cloning to combine both acoustic and linguistic validity. Using only 20 minutes of electrocorticography (ECoG) data per subject, our approach achieves highly intelligible synthesized speech (mean opinion score = 4.0/5.0, word error rate = 18.9%). Our dual-path framework reconstructs natural and intelligible speech from ECoG, resolving the acoustic-linguistic trade-off.

Introduction

Understanding how the human brain encodes and decodes spoken language is a central challenge in cognitive neuroscience (1–5), with profound implications for brain-computer interfaces (BCIs) (6–9). Reconstructing perceived speech directly from neural activity provides a powerful approach to studying the hierarchical organization of language processing in the auditory cortex and to developing assistive technologies for individuals with impaired communication abilities (10–15). This neural decoding paradigm not only helps reveal the representational structure of linguistic features across the cortex but also enables the synthesis of intelligible speech directly from brain signals (16–18).

Recent advances in deep learning have accelerated progress in neural speech reconstruction by enabling end-to-end mappings from neural recordings to acoustic waveforms (6, 7, 10, 14, 19). However, these models typically require large-scale paired datasets—hours or even days of simultaneous neural and speech recordings per subject—to learn the complex nonlinear transformations involved. This data scarcity remains a major bottleneck, especially in clinical or intracranial recording settings where data acquisition is constrained by practical and ethical limitations.

To address this challenge, recent work has leveraged pre-trained self-supervised learning (SSL) models, such as Wav2Vec2.0 and HuBERT, which learn rich acoustic and phonetic representations from large-scale unlabeled speech corpora (20, 21). These models have been shown to exhibit strong correspondence with neural activity in the auditory cortex, supporting near-linear mappings between latent model representations and cortical responses (22–24). Similarly, large language models trained purely on text, such as GPT (25), also capture high-level semantic representations that correlate with activity in language-responsive cortical regions (26–28). Together, these findings suggest that the latent spaces of speech and language models provide a natural bridge between brain activity and interpretable representations of speech, enabling more data-efficient neural decoding.

Current approaches to speech decoding from brain activity can be broadly categorized into two paradigms. The first, neural-to-acoustic decoding, maps brain signals to low-level spectral or waveform features such as formants, pitch, or spectrograms (10, 11, 15, 29). This approach captures fine acoustic detail, including speaker identity, prosody, and timbre, and benefits from pre-trained speech synthesizers for waveform generation (17–19, 30, 31). However, due to the high dimensionality and dynamic nature of speech, this regression problem is inherently difficult and often yields low intelligibility when training data is limited. The second, neural-to-text decoding, treats speech reconstruction as a temporal classification task—mapping neural signals to a sequence of discrete linguistic units such as phonemes, characters, or words (12, 14, 32–34). This strategy achieves strong performance with relatively small datasets and aligns well with natural language processing (NLP) pipelines (35–38), but sacrifices naturalness and expressiveness and lacks paralinguistic features crucial for human communication.

Here, we propose a unified and efficient dual-pathway decoding framework that integrates the complementary strengths of both paradigms. Our method maps intracranial electrocorticography (ECoG) signals into the latent spaces of pre-trained speech and language models via two lightweight neural adaptors: an acoustic pathway, which captures low-level spectral features for naturalistic speech synthesis, and a linguistic pathway, which extracts high-level linguistic tokens for semantic intelligibility. These pathways are fused using a fine-tuned text-to-speech (TTS) generator with voice cloning, producing re-synthesized speech that retains both the acoustic spectrotemporal details, such as the speaker’s timbre and prosody, and the message linguistic content. The adaptors rely on near-linear mappings and require only 20 minutes of neural data per participant for training, while the generative modules are pre-trained on large unlabeled corpora and require no neural supervision.

Results

Overview of the proposed speech re-synthesis method

Our proposed framework for reconstructing speech from intracranial neural recordings is designed around two complementary decoding pathways: an acoustic pathway focused on preserving low-level spectral and prosodic detail, and a linguistic pathway focused on decoding high-level textual and semantic content. These two streams are ultimately fused to synthesize speech that is both natural-sounding and intelligible, capturing the full richness of spokexn language. Fig. 1 provides a schematic overview of this dual-pathway architecture.

The neural data acquisition and acoustic-linguistic dual-pathway framework for neural-driven natural speech re-synthesis.

(A) The neural data acquisition. We collected ECoG data from 9 monolingual native English participants as they each listened to English sentences from the TIMIT corpus, resulting in 20 minutes of neural recording data per participant. Furthermore, for each participant, 70%, 20%, and 10% of the recorded neural activities are respectively allocated to the training, validation, and test sets randomly. (B) The acoustic-linguistic dual-pathway framework. The acoustic pathway consists of two stages: Stage 1: A HiFi-GAN generator is pre-trained to synthesize natural speech from features extracted by the frozen Wav2Vec2.0 encoder, using a multi-receptive field fusion module and adversarial training with discriminators. This stage uses the LibriSpeech corpus to enhance speech representation learning. Stage 2: A lightweight LSTM adaptor maps neural activity to speech representations, enabling the frozen HiFi-GAN generator to re-synthesize high acoustic fidelity speech from neural data. The linguistic pathway involves a Transformer adaptor refining neural features to align with word tokens, which are fed into the frozen Parler-TTS model to generate high linguistic intelligibility speech. The voice cloning stage uses CosyVoice 2.0 (fine-tuned on TIMIT) to clone the speaker’s voice, ensuring the ultimate re-synthesized speech waveform matches the original stimuli in clarity and voice characteristics.

The acoustic pathway aims to synthesize speech with high acoustic naturalness, including prosody, speaker identity, and timbre, through a two-stage training process. In stage 1, we pre-train a high-fidelity generative adversarial network (HiFi-GAN) generator using a large corpus of natural speech data, without requiring paired neural recordings. This stage involves a speech autoencoder architecture consisting of a frozen Wav2Vec2.0 encoder (20), which transforms ECoG signals into latent representations known to correlate with brain activity (22, 23), and a trainable HiFi-GAN decoder, which synthesizes natural-sounding speech waveforms from these representations (Fig. 1B). HiFi-GAN (39), a generative adversarial network, was pre-trained to reconstruct high-fidelity speech using features extracted by the Wav2Vec2.0 encoder. This stage defines a rich, brain-aligned acoustic feature space (24) that does not require paired neural data. In stage 2, in order to decode speech from the brain, we next trained a lightweight acoustic adaptor that maps ECoG recordings onto the pre-defined latent space of the speech generator. This adaptor, a three-layer bidirectional long-short term memory (LSTM) model (~9.4M parameters), was trained using only 20 minutes of neural recordings per participant while listening to natural speech. The HiFi-GAN waveform decoder remained frozen, allowing the adaptor to be trained efficiently and independently. This acoustic pathway enables high-fidelity speech reconstruction directly from neural signals, capturing fine-grained auditory detail while remaining data-efficient.

While the acoustic pathway ensures naturalness, it lacks explicit linguistic structure. The goal of the linguistic pathway is to decode discrete word-level representations from brain activity, facilitating the reconstruction of speech that is not only fluent but also semantically and syntactically accurate. To extract high-level linguistic information, we trained a Transformer-based linguistic adaptor (~10.1M parameters) to align ECoG signals with word token representations derived from transcribed speech. This adaptor comprises a three-layer encoder and decoder and is trained in a supervised manner using the same 20-minute neural dataset. The adaptor output is fed into Parler-TTS (40), a state-of-the-art TTS model (~880M parameters), which synthesizes fluent speech from word token sequences. This pathway captures syntactic structure and lexical identity that are not explicitly represented in the acoustic stream.

To generate speech that combines the strengths of both pathways, retaining the acoustic richness from the first and the intelligibility from the second, we fused the outputs in a final synthesis stage. The resulting waveforms were passed through CosyVoice 2.0 (41), a high-quality speech synthesis model with robust voice cloning and denoising capabilities.

CosyVoice 2.0 was fine-tuned on the training set of TIMIT to improve articulation clarity, speaker-specific characteristics, and prosodic expressiveness. The final output preserves both low-level acoustic fidelity (e.g., pitch, prosody, and timbre) and high-level linguistic content (e.g., accurate word sequences), resolving the traditional trade-off in brain-to-speech decoding. Across the entire framework, unsupervised pretraining and supervised neural adaptation are used synergistically to achieve efficient, high-quality speech reconstruction.

The acoustic and linguistic performance of the reconstructed speech

We collected intracranial neural recordings from nine participants (4 male and 5 females; age range: 31-55) undergoing clinical neurosurgical evaluation for epilepsy. Each participant was implanted with high-density ECoG electrode arrays placed over the superior temporal gyrus and surrounding auditory speech cortex, based on clinical needs. During the experiment, participants passively listened to a set of 599 naturalistic English sentences spoken by multiple speakers. Neural signals were recorded at high temporal resolution while preserving the spatial specificity necessary to capture cortical responses to speech. This yielded approximately 20 minutes of labelled neural data for each participant. We then evaluated the quality of the re-synthesized sentences for each participant in the test set. Both subjective and objective evaluation metrics were adopted to give a comprehensive quantification of the effectiveness of this dual-path strategy in reconstructing intelligible and naturalistic speech from limited neural data. This included mean opinion score (MOS), mel-spectrogram correlation, word error rate (WER), and phoneme error rate (PER).

First, we evaluated the lower-level acoustic quality of the reconstructed speech. The average mel-spectrogram R2 (Fig. 2B) for the neural-driven re-synthesized speech across participants was 0.824 ± 0.029 (mean ± s.e., the best performance across participants achieving 0.844 ± 0.028, Fig. S3B). For comparison, we also computed these metrics on surrogate speech with additive noise (42) (measured in dB). The quality of the reconstructed speech was comparable to the R2 of −5dB and 0dB additive noise (−5dB: mel-spectrogram R2 = 0.864 ± 0.019, p = 0.161, two-sided t-test; 0dB: mel-spectrogram R2 = 0.771 ± 0.014, p = 0.064, two-sided t-test). The average MOS (Fig. 2c) for the neural-driven re-synthesized speech across participants was 3.956 ± 0.173, with the best performance across participants reaching 4.200 ± 0.119 (Fig. S3C).

Examples of neural-driven speech reconstruction performance.

(A) Ground truth and decoded speech comparisons. Waveforms (time vs. amplitude) and mel-spectrograms (time vs. frequency range 0-8 kHz) are shown for both the original (top) and reconstructed (bottom) speech samples, demonstrating preserved spectral-temporal patterns in the neural-decoded output. Decoding demonstration of phonemes and words is attached to the speech. (B) Mel-spectrogram correlation analysis. The KDE curve shows the distribution of aligned correlation coefficients between the original and reconstructed mel-spectrograms across all test samples, while the bar chart represents the percentage of trials in the test set (mean = 0.824 ± 0.028). Higher values reflect better acoustic feature preservation in the time-frequency domain. Purple dashed lines (light to dark) show average R2 after adding −10 dB, −5 dB, and 0 dB additive noise to the original speech. (C) Human subject evaluations. The KDE curve displays the distribution of human evaluation results, while the bar chart represents the percentage of trials in the test set (mean = 3.956 ± 0.175). Higher values reflect better intelligibility. Purple dashed lines (light to dark) show the average MOS after adding −20 dB, −10 dB, and 0 dB additive noise to the original speech. (D-E) Word Error Rate (WER) and Phoneme Error Rate (PER) assessment.

Second, we further evaluated the linguistic content and intelligibility of the re-synthesized speech using word error rate (WER) and phoneme error rate (PER) metrics (Fig. 2D, E). The neural-driven re-synthesized speech achieved an average WER of 18.9 ± 3.3% across participants, with the best performance reaching 12.1 ± 2.1% (Fig. S3D). The WER result is comparable to adding −5dB noise on the original input speech stimuli (14.0 ± 2.1%, p = 0.332, two-sided t-test). Similarly, the average PER was 12.0 ± 2.5%, with the best performance reaching 8.3 ± 1.5% (Fig. S3E), comparable to adding −5dB noise on the original input speech stimuli (6.0 ± 1.5%, p = 0.068, two-sided t-test). evaluation. The KDE curve displays distribution, while the bar chart represents the percentage of trials in the test set (mean WER = 0.189 ± 0.033, mean PER = 0.120 ± 0.025). Lower values indicate better word-level reconstruction accuracy. Purple dashed lines (light to dark) show average WER after adding −10 dB, −5 dB, and 0 dB additive noise to the original speech.

The acoustic pathway reconstructs high-fidelity lower-level acoustic information

The acoustic pathway, implemented through a bi-directional LSTM neural adaptor architecture (Fig. 1B), specializes in reconstructing fundamental acoustic properties of speech. This module directly processes neural recordings to generate precise time-frequency representations, focusing on preserving speaker-specific spectral characteristics like formant structures, harmonic patterns, and spectral envelope details. Quantitative evaluation confirms its core competency: achieving a mel-spectrogram R2 of 0.793 ± 0.016 (Fig. 2B) demonstrates remarkable fidelity in reconstructing acoustic microstructure. This performance level is statistically indistinguishable from original speech degraded by 0dB additive noise (0.771 ± 0.014, p = 0.242, one-sided t-test).

However, exclusive reliance on acoustic reconstruction reveals fundamental limitations. Despite excellent spectral fidelity, the pathway produces critically impaired linguistic intelligibility. At the word level, intelligibility remains unacceptably low (WER = 74.6 ± 5.5%, Fig. 2D), while MOS and phoneme-level precision fares only marginally better (MOS = 2.878 ± 0.205, Fig. 2C; PER = 28.1 ± 2.2%, Fig. 2E). These deficiencies manifest perceptually as distorted articulation, inconsistent prosody, and compromised phonetic distinctiveness - collectively rendering the output functionally unintelligible for practical communication. This performance gap underscores a fundamental constraint: accurate acoustic feature reconstruction alone is insufficient for generating comprehensible speech, necessitating complementary linguistic processing.

The linguistic pathway reconstructs high-intelligibility, higher-level linguistic information

The linguistic pathway, instantiated through a pre-trained TTS generator (Fig. 1B), excels in reconstructing abstract linguistic representations. This module operates at the phonological and lexical levels, converting discrete word tokens into continuous speech signals while preserving prosodic contours, syllable boundaries, and phonetic sequences. It achieves a mean opinion score of 4.822 ± 0.086 (Fig. 3C) - significantly surpassing even the original human speech (4.234 ± 0.097, p = 6.674×10−33). Complementing this perceptual quality, objective intelligibility metrics confirm outstanding performance: WER reaches 17.7 ± 3.2%, with PER at 11.0 ± 2.3%. These results validate the pathway’s capacity for reconstructing high-level linguistic structures from symbolic representations.

Performance comparison among re-synthesized speech and MLP-regression, acoustic and linguistic baselines.

(A) Speech waveform and mel-spectrogram observations. Depicted are waveforms (time vs. amplitude) and mel-spectrograms (time vs. frequency ranging 0-8 kHz) for illustrative speech samples (Ground truth, MLP regression and acoustic-linguistic pathway intermediate outputs as baseline 1-3, and our ultimate re-synthesized natural speech). (B) Objective evaluation using mel-spectrogram R2. Violin plots show the aggregated distribution of R2 scores (0-1 scale) assessing spectral fidelity. The white dot represents the median, the box spans the interquartile range, and whiskers extend to ±1.5×IQR. Three dashed lines (light to dark) show average R2 after adding −10 dB, −5 dB, and 0 dB additive noise to the original speech. (C) Subjective quality evaluation using mean opinion score (MOS). Violin plots show aggregated MOS distribution (1-5 scale). Three dashed lines (light to dark) indicate average MOS after adding −20 dB, −10 dB, and 0 dB additive noise. GT: Ground truth. (D) Intelligibility assessment using word error rate (WER). Violin plots show aggregated WER distribution (0-1 scale). Three purple dashed lines (light to dark) show the average WERs after adding −10 dB, −5 dB, and 0 dB additive noise. (E) Phoneme error rate (PER) assessment. Format and noise conditions identical to panel D. Statistical significance markers: *:p<0.01, *:*p<0.001, n.s.: not significant.

Nevertheless, this linguistic feature caption comes at an acoustic cost. The pathway exhibits significant limitations in spectral fidelity, achieving only 0.627 ± 0.026 (Fig. 3B) mel-spectrogram R2 - substantially below the 0dB noise reference (0.771 ± 0.014, p = 2.919×10−8). This deficiency stems from fundamental architectural constraints: lacking explicit acoustic feature modelling, the system fails to preserve critical speaker-specific attributes like vocal tract resonances, glottal source characteristics, and individual articulatory patterns. Consequently, while linguistically intelligible, the output lacks authentic speaker identity and natural timbral qualities essential for realistic speech reconstruction.

The fine-tuned voice cloner further enhances the fidelity and intelligibility of re-synthesized speech

Voice cloning is used to bridge the gap between acoustic fidelity and linguistic intelligibility in speech reconstruction. This approach strategically combines the strengths of complementary pathways: the acoustic pathway preserves speaker-specific spectral characteristics while the linguistic pathway maintains lexical and phonetic precision. By integrating these components through neural voice cloning, we achieve balanced reconstruction that overcomes the limitations inherent in isolated systems.

The integration paradigm resolves this fundamental dichotomy through strategic combination. Where the acoustic pathway preserves spectral patterns (R2 = 0.793 ± 0.016) but suffers unintelligibility (WER = 74.6 ± 5.5%), and the linguistic pathway delivers lexical precision but lacks acoustic fidelity, their fusion creates a complementary system. This synergistic architecture maintains the linguistic pathway’s exceptional intelligibility while incorporating the acoustic pathway’s spectrotemporal reconstruction. The resultant output achieves what neither can accomplish independently: linguistically precise speech with an authentic acoustic signature, effectively decoupling and optimizing both reconstruction dimensions within a unified framework.

To validate this integrated approach, we established three baseline models for comprehensive benchmarking. Baseline 1 (MLP-regression) directly maps neural recordings to mel-spectrograms without specialized processing. Baseline 2 represents the output of our acoustic pathway implementation, optimized for spectral feature reconstruction. Baseline 3 corresponds to the linguistic pathway output, generated by the TTS system without speaker-specific timbre adaptation.

The integration breakthrough occurs when combining this acoustic foundation with linguistic pathway components. The full pipeline achieves transformative improvements: WER decreases by 75% (18.9 ± 3.3% vs 74.6 ± 5.5%, p = 4.249×10−88, one-sided t-test, Fig. 2D), while PER drops by 58% (12.0 ± 2.5% vs 28.1 ± 2.2%, p = 2.592×10−44, one-sided t-test, Fig. 2E). Crucially, this intelligibility enhancement occurs without compromising acoustic fidelity: Mel spectrum R2 = 0.824±0.029 vs 0.793 ± 0.016 (acoustic pathway, p = 4.486×10−3, one-sided t-test).

To further elucidate the phoneme recognition performance, we conducted an in-depth analysis of the errors present in the transcribed phoneme sequences derived from both original and re-synthesized speech. Specifically, we focused on the re-synthesized speech, where individual characters were subjected to insertion, deletion, or substitution operations to align them with the ground truth. This process not only quantified the errors but also provided insights into the relationships between different phonemes.

The confusion matrices (Fig. 4 A-D) illustrate the distribution of these errors. The diagonal elements represent the correct matches, indicating the accuracy for each phoneme, while off-diagonal non-zero elements denote substitution errors. Empty rows signify insertion errors (extraneous phonemes), and empty columns indicate deletion errors (missing phonemes). These visual representations highlight the specific challenges faced by our proposed model and the baseline models in accurately recognizing certain phonemes.

Phoneme recognition performance: confusion matrices and accuracy comparison

(A) Confusion matrix between transcribed phoneme sequences using our proposed model (horizontal axis) and ground truth (vertical axis). Diagonal values represent recognition accuracy for each phoneme (correct matches), while off-diagonal non-zero elements indicate substitution errors. Empty rows denote insertion errors (extraneous phonemes), empty columns indicate deletion errors (missing phonemes). All non-zero elements are highlighted for visual clarity. (B-D) Phoneme confusion matrix using baseline 1-3 for phoneme transcription. (E) Phoneme class clarity across methods. The violin plots show the proportion of correctly identified phonemes for each symbol. Asterisks (*) indicate phonemes where our proposed framework significantly outperforms baselines.

In addition to the confusion matrices, we categorized the phonemes into vowels and consonants to assess the phoneme class clarity. We defined “phoneme class clarity” (PCC) as the proportion of errors where a phoneme was misclassified within the same class versus being misclassified into a different class. For instance, a vowel being mistaken for another vowel would be considered a within-class error, whereas a vowel being mistaken for a consonant would be classified as a between-class error.

Our analysis revealed that the re-synthesized speech achieved a phoneme class clarity of 2.462 ± 0.201 (Fig. 4E), significantly outperforms chance level (1.360, calculated under the assumption that substituted phonemes are randomly distributed among all other phonemes, p = 3.424×10−6, one-sided t-test), MLP regression and acoustic pathway output (MLP regression: PCC = 1.694 ± 0.134, p = 1.312×10−3, one-sided t-test; acoustic pathway output: PCC = 1.735 ± 0.140, p = 2.367×10−3; linguistic pathway output: PCC = 2.446 ± 0.203, p = 0.477, one-sided t-test), indicating that most errors occurred within the same phoneme class. This result suggests that the model has a relatively good understanding of the phonetic structure, as it tends to confuse similar sounds rather than completely unrelated ones. Notably, this metric did not show significant differences compared to the baselines, suggesting that all models exhibit similar tendencies in terms of phoneme class confusion.

These comprehensive results confirm that our voice cloning framework successfully integrates the complementary strengths of acoustic and linguistic processing pathways. It achieves this integration without compromising either dimension: while significantly enhancing acoustic fidelity beyond what the linguistic pathway alone can achieve, it simultaneously maintains near-optimal linguistic intelligibility that the isolated acoustic pathway cannot provide. This balanced performance profile demonstrates the efficacy of our approach in producing high-quality, intelligible speaker-specific speech reconstruction.

Discussion

In this study, we present a dual-path framework that synergistically decodes both acoustic and linguistic speech representations from ECoG signals, followed by a fine-tuned zero-shot text-to-speech network to re-synthesize natural speech with unprecedented fidelity and intelligibility. Crucially, by integrating large pre-trained generative models into our acoustic reconstruction pipeline and applying voice cloning technology, our approach preserves acoustic richness while significantly enhancing linguistic intelligibility beyond conventional methods. Using merely 20 minutes of ECoG recordings, our model achieved superior performance with a WER of 18.9% ± 3.3% and PER of 12.0% ± 2.5% (Fig. 2D, E). This integrated architecture, combining pre-trained acoustic (Wav2Vec2.0 and HiFi-GAN) and linguistic (Parler-TTS) models through lightweight neural adaptors, enables efficient mapping of ECoG signals to dual latent spaces. Such methodology substantially reduces the need for extensive neural training data while achieving breakthrough word clarity under severe data constraints. The results demonstrate the feasibility of transferring the knowledge embedded in speech-data pre-trained artificial intelligence (AI) models into neural signal decoding, paving the way for more advanced brain-computer interfaces and neuroprosthetics.

Our framework establishes a framework for speech decoding by outperforming prior acoustic-only or linguistic-only approaches through integrated pretraining-powered acoustic and linguistic decoding. This dual-path methodology proves particularly effective where traditional methods fail, as it simultaneously resolves the acoustic-linguistic trade-off that has long constrained reconstruction quality. While end-to-end re-synthesis remains intuitively appealing, prior work confirms that direct methods achieve only modest performance given neural data scarcity(15). To overcome this, we propose a hybrid encoder-decoder architecture utilizing: (1) pre-trained spectral synthesizers (Wav2Vec2.0 and HiFi-GAN) for acoustic fidelity, and (2) transformer-based token decoders (Parler-TTS) for linguistic precision. Participant-specific projection modules further ensure cross-subject transferability with minimal per-user calibration. Collectively, these advances surmount core limitations of direct decoding, enabling unprecedented speech quality within extreme data constraints.

A pivotal advancement lies in establishing robust, clinically relevant intelligibility metrics, notably achieving 18.9% WER and 12.0% PER through standardized evaluation, directly addressing the core challenge of word recognition in speech decoding. Our dual-path framework enables comprehensive sentence-level assessment through objective benchmarks (phonetic precision via PER and lexical accuracy via WER evaluated by Whisper ASR), acoustic fidelity validation (high mel-spectrogram correlation: 0.824 ± 0.029, Fig. 2B) and human perceptual testing (near-’excellent’ MOS ratings: 3.956 ± 0.173, Fig. 2C). Critically, this tripartite evaluation spanning acoustic (spectral/time-domain), linguistic (phoneme/word), and perceptual dimensions revealed superior reconstruction quality, while objective metrics confirmed spectral coherence rivalling clean speech inputs.

Our findings demonstrate that neural representations exhibit dual alignment, which is not only with acoustic features from deep speech models (Wav2Vec2.0), but critically with linguistic features from language models (Parler-TTS), establishing a bidirectional bridge between cortical activity and hierarchical speech representations. This convergence mirrors the complementary processing streams observed in the human speech cortex (23, 27) where self-supervised models capture both spectrotemporal patterns and semantic structures. Such unified alignment marks a paradigm shift in brain-AI integration: The discovery of near-linear mappings between neural and multimodal AI spaces unlocks transformative applications in speech synthesis and cognitive interfaces. Furthermore, our results confirm that foundation models serve as scalable ‘cognitive mirrors’ (43). With the advent of more sophisticated generative models, we anticipate enhanced neural decoding capabilities, including the potential to improve signal quality (in terms of SNR) and refine the generative model itself. Additionally, this framework extends beyond speech to other modalities, such as vision, suggesting that similar principles may apply to the generation of visual content from neural signals.

This work advances speech decoding and potentially speech BCIs by enabling real-time speech reconstruction with minimal neural data, improving scalability and practicality through pre-trained AI models, and suggesting potential for broader applications in perception, memory, and emotion. By adapting neural activity to a common latent space of the pre-trained speech autoencoder, our framework significantly reduces the quantity of neural data required for effective speech decoding, thereby addressing a major limitation in current BCI technologies. This reduction in data needs paves the way for more accessible and scalable BCI solutions, particularly for individuals with speech impairments who stand to benefit from immediate and intelligible speech reconstruction. Furthermore, the applicability of our model extends beyond speech, hinting at the possibility of decoding other cognitive functions once the corresponding neural correlates are identified. This opens up exciting avenues for expanding BCI functionality into areas such as perception, memory, and emotional expression, thereby enhancing the overall quality of life for users of neuroprosthetic devices.

There are several limitations in our study. The quality of the re-synthesized speech heavily relies on the performance of the generative model, indicating that future work should focus on refining and enhancing these models. Currently, our study utilized English speech sentences as input stimuli, and the performance of the system in other languages remains to be evaluated. Regarding signal modality and experimental methods, the clinical setting restricts us to collecting data during brief periods of awake neurosurgeries, which limits the amount of usable neural activity recordings. Overcoming this time constraint could facilitate the acquisition of larger datasets, thereby contributing to the re-synthesis of higher-quality natural speech. Additionally, exploring non-invasive methods represents another frontier; with the accumulation of more data and the development of more powerful generative models, it may become feasible to achieve effective non-invasive neural decoding for speech re-synthesis.

In summary, our dual-path framework achieves high speech reconstruction quality by strategically integrating language models for lexical precision and voice cloning for vocal identity preservation—yielding a 37.4% improvement in MOS scores over conventional methods. This approach enables high-fidelity, sentence-level speech synthesis directly from cortical recordings while maintaining speaker-specific vocal characteristics. Despite current constraints in generative model dependency and intraoperative data collection, our work establishes a new foundation for neural decoding development. Future efforts should prioritize: (1) refining few-shot adaptation techniques, (2) developing non-invasive implementations, and (3) expanding to dynamic dialogue contexts. The convergence of neurophysiological data with multimodal foundation models promises transformative advances—not only revolutionizing speech BCIs but potentially extending to cognitive prosthetics for memory augmentation and emotional communication. Ultimately, this paradigm will deepen our understanding of neural speech processing while creating clinically viable communication solutions for those with severe speech impairments.

Materials and Methods

Participants

The study comprised 9 monolingual right-handed participants, each with electrodes placed on the left hemisphere cortex to clinically monitor seizure activities. Electrode placement followed specific clinical requirements (see Fig. S1 for grid placement in each placement). Participants were thoroughly informed about the experiment, as outlined in the consent document approved by the Institutional Review Board (IRB) of the University of California, San Francisco. They volunteered for participation, ensuring that their involvement did not influence their clinical care. Additionally, verbal consent was obtained from participants at the beginning of each experimental session.

Data acquisition and neural signal processing

All participants utilized high-density ECoG grids of uniform specifications and type. During the experimental tasks, neural signals captured by the ECoG grids were acquired using a multi-channel amplifier optically linked to the digital signal processor. Data recording was performed through Tucker-Davis Technology (TDT) OpenEx software. The local field potential at each electrode contact was amplified and sampled at 3052 Hz. Subsequent to data collection, the experiment employed the Hilbert transform to compute the analytic amplitudes of eight Gaussian filters (with center frequencies ranging from 70 to 150 Hz), and the signal was down-sampled to 50 Hz. These tasks were segmented into record blocks lasting approximately 5 minutes each. The neural signals were z-score normalized for each recording block.

Experimental Stimuli

The acoustic stimuli employed in this study comprised natural and continuous English speech. The English speech stimuli were sourced from the TIMIT corpus (44), consisting of 499 sentences read by 286 male and 116 female speakers. A silence interval of 0.4 seconds was maintained between sentences. The task was organized into five blocks, each with a duration of approximately 5 minutes.

Electrode localization

In cases involving chronic monitoring, the electrode placement procedure involved pre-implantation MRI and post-implantation CT scans. In awake cases, temporary high-density electrode grids were utilized to capture cortical local potentials, with their positions recorded using the Medtronic neuronavigation system. The recorded positions were then aligned to pre-surgery MRI data, with intraoperative photographs serving as additional references. Localization of the remaining electrodes was achieved through interpolation and extrapolation techniques.

Speech-responsive electrode selection

The selection of speech-responsive electrodes played a crucial role in the re-synthesis of high-quality natural speech and in avoiding overfitting. Onsets were identified as the initiation of speech and are preceded by more than 400 ms of silence. A paired sample t-test was conducted to investigate whether the average post-onset response (400 ms to 600 ms) exhibited a significant increase compared to the average pre-onset response (−200 ms to 0 ms, p < 0.01, one-sided, Bonferroni corrected).

Architecture and training of the speech Autoencoder

The speech Autoencoder served as a pivotal component, taking natural speech as input and reproducing the input itself. Within this model, the intermediate encoding layers were construed as the feature extraction layers. The architecture comprises an encoder, namely Wav2Vec2.0, and a decoder, which is the HiFi-GAN generator.

The Wav2Vec2.0 encoder was comprised of 7 1D convolutional layers that down-sample 16kHz natural speech to 50Hz, thereby extracting 768-dimensional local features. Additionally, it integrated 12 transformer-encoder blocks to extract contextual feature representations from unlabeled speech data.

On the other hand, the HiFi-GAN generator incorporated 7 transposed convolutional layers for up-sampling, along with multi-receptive field (MRF) fusion modules. Each MRF module aggregated the output features from multiple ResBlocks. The discriminator within the HiFi-GAN encompassed both multi-scale and multi-period discriminators. Notably, the up-sample rates and kernel sizes of each ResBlock in the MRFs were meticulously configured. The up-sample rates of each ResBlock in the MRFs were set as 2, 2, 2, 2, 2, 2, 5 respectively. The up-sample kernel sizes of each ResBlock in the MRFs were set as 2, 2, 3, 3, 3, 3, 10. The periods of the multi-period discriminator were set as 2, 3, 5, 7, 11.

The HiFi-GAN generator G and discriminators Dk were trained simultaneously and adversarially. The loss function consisted of three parts: adversarial loss LAdv to train the HiFi-GAN, mel-spectrogram loss LMel to improve the training efficiency of the generator and feature matching loss LFM to measure the similarity between original speech and re-synthesized speech from learned features: The regularization coefficients λFM and λMel were set as 2 and 50 respectively in the implementation. In detail, the three terms of LG were: Where φ represented the mel-spectrogram operator, T was the number of layers in the discriminator, Di and Ni were the features and the number of features in the ith layer of the discriminator, x and xSAE represented the original speech and the reconstructed speech using the pre-trained speech Autoencoder: The parameters in Wav2Vec2.0 were frozen within this training phase. The parameters in HiFi-GAN were optimized using the Adam optimizer with a fixed learning rate of 10−5, β1 = 0.9, β2 = 0.999. We trained this Autoencoder in LibriSpeech, a 960-hour English speech corpus with a sampling rate of 16kHz. We spent 12 days in parallel training on 6 Nvidia GeForce RTX3090 GPUs. The maximum training epoch was 2000. The optimization did not stop until the validation loss no longer decreased.

Acoustic feature adaptor training and ablation test

The acoustic feature adaptor was trained to acquire the acoustically fidelity re-synthesized natural speech. The acoustic speech re-synthesizer received z-scored high gamma ECoG snippets with dimensions N × T, where N was the number of selected speech-responsive electrodes, and T was the ECoG recordings (50 Hz) down-sampled to the same sample rate as the speech features extracted from pre-trained Wav2Vec2.0.

A three-layer bidirectional LSTM adaptor f (see Fig. 1B) aligned recorded ECoG signals to speech features. The output of the adaptor was 768 × T, as the input of the HiFi-GAN generator pre-trained in phase 1. The ultimate output of the HiFi-GAN was a speech waveform with a length of 1 × (320T + 80) in a 16K sample rate.

We froze all the parameters of the pre-trained speech Autoencoder while training the adaptor. The loss function used in this phase consisted of two parts: the mel-spectrogram loss LMel and the Laplacian loss LLap: where the regularization coefficient λ was set as 0.1 using the ablation test.

LMel was used to enhance the acoustic fidelity and intelligibility of the ultimate re-synthesized speech from recorded neural activity: where x and xa respectively represented the original and acoustic fidelity speech, and Φ represented the mel-spectrogram operator.

In order to improve the phonetic intelligibility, a Laplace operator as a convolution kernel was applied to the mel-spectrogram, which was a simplified representation of the convolution on the spectrogram, while convolution on the spectrogram was proven to be effective on speech denoising and enhancement(45): The parameters in the HiFi-GAN generator were frozen within this training phase. The parameters in the adaptor were trained in Stochastic Gradient Descent (SGD) optimizer with a fixed learning rate of 3 × 10−3 and a momentum of 0.5. A 10% dropout layer was added in this stage to avoid overfitting. The optimization did not stop until the validation loss no longer decreased.

For each participant, we trained a personalized feature adaptor. Each individual feature adapter underwent training for 500 epochs on one NVidia GeForce RTX3090 GPU, which required 12 hours to complete. The dataset was divided into training, validation, and test sets, comprising 70%, 20%, and 10% of the total trials, respectively.

Linguistic feature adaptor training and ablation test

The linguistic feature adaptor was designed to decode word token sequences from ECoG recordings. The model received z-scored high gamma ECoG snippets with dimensions (N × T), where N was the number of selected speech-responsive electrodes, and T was the ECoG recordings (50 Hz) down-sampled to the same sample rate as the acoustic feature training stage.

An attention-based Seq2Seq Transformer architecture (see Fig. 1B) was employed to adapt ECoG signals to linguistic features. The linguistic adaptor consisted of 3 encoder layers and 3 decoder layers, with a hidden dimension of 256 and 8 attention heads. The input ECoG signals were first projected into a latent space using a linear layer, followed by positional encoding to capture temporal dependencies. The decoder generated word token sequences autoregressively, starting with a start-of-sequence (SOS) token. The output dimension matched the linguistic feature space (1024-dimensional).

The loss function combined the token-level KL-divergence and sequence length prediction loss with L2 regularization: where Ltoken measured the KL-divergence between predicted and target token distributions: and Llength was a Huber loss between predicted and actual sequence lengths: The regularization coefficient λ1 was set to 1 based on ablation tests, and L2 regularization (weight decay = 0.01) was applied to mitigate overfitting. The model was trained using the Adam optimizer with a fixed learning rate of 10−4 for 500 epochs. The dataset was split into training (70%), validation (20%), and test (10%) sets, consistent with the acoustic feature adaptor. For each participant, a personalized linguistic adaptor was trained on one NVidia GeForce RTX3090 GPU, requiring approximately 6 hours to complete.

Automatic phoneme recognition of re-synthesized speech sentences

In order to quantitatively evaluate the synthesized speech is to recognize the phonemes, we fine-tuned a phoneme recognizer using Wav2Vec2.0, a logistic regression phoneme classifier, and a phoneme decoder. Wav2Vec2.0 was used for feature extraction, the classifier outputs a probability vector from Wav2Vec2.0 features, and the phoneme decoder decoded a phoneme sequence from the probability vectors to finally get the phoneme sequence.

The models are trained and evaluated on the training set and test set of TIMIT, and tested on the re-synthesized ECoG recordings. These parameters were trained using the AdamW optimizer with a fixed learning rate of 10−4, β1 = 0.9, β2 = 0.999, and ε = 10−8 in the CTC loss function. The optimization did not stop until the validation loss no longer decreased.

Phoneme error rate (PER) and word error rate (WER)

We implemented a two-stage transcription pipeline to evaluate linguistic intelligibility. First, word sequences were transcribed using OpenAI’s Whisper model(46) (base version) with default parameters, which provided both word-level alignments and confidence scores. The reference and decoded word sequences were then aligned using edit distance to compute WER. The error rates were calculated by applying the same edit distance algorithm to the forced aligned phoneme sequences. This combined approach ensured consistent phoneme representations between natural and synthesized speech during comparison: Where Edit(original, transcribed) denotes the edit distance between original and transcribed phoneme sequences; N is the number of phonemes in the sentence.

Alignment of the ultimate re-synthesized speech to calculate the mel-spectrogram R2

To quantitatively evaluate the acoustic fidelity of the re-synthesized speech using mel-spectrogram R2, we performed temporal alignment between the reconstructed and ground truth mel-spectrograms. First, word-level timestamps were extracted from the transcribed speech using OpenAI’s Whisper model, which provided precise boundaries for voiced segments. The mel-spectrogram of the re-synthesized speech was then segmented into these intervals, excluding silent or non-speech regions. Each voiced segment was resampled to match the temporal resolution of the corresponding ground truth spectrogram using linear interpolation, ensuring frame-by-frame comparability. The squared Pearson correlation coefficient (R2) was computed over the aligned spectrograms, measuring the proportion of variance in the ground truth spectrogram explained by the reconstructed features. This approach accounted for natural variations in speech rate while rigorously quantifying spectral fidelity, with higher R2 values (closer to 1) indicating superior preservation of time-frequency structures critical for speech intelligibility and naturalness.

Voice cloning for the ultimate re-synthesized natural speech

To achieve personalized speech output, we implemented a voice cloning pipeline that preserves the subject’s unique vocal features (Fig. S2). The system takes two key inputs: (1) a clean voice sample from the denoised linguistic speech reference, and (2) the corresponding word sequence transcribed by Whisper. This approach serves dual purposes - Whisper first transcribes the original speech to provide accurate text prompts, then later evaluates the intelligibility of the ultimate re-synthesized speech by transcribing it again for word error rate calculation.

The cloning process begins with extracting vocal fingerprints from the short reference sample, capturing distinctive timbral qualities like pitch contour and formant structure. These vocal features are then mapped to the target word sequences, generating synthetic speech that maintains the subject’s natural voice patterns while clearly articulating the desired content.

We specifically employed the “3-second rapid cloning” mode to optimize for both efficiency and voice quality. This ultra-fast processing allows for quick adaptation to individual speakers while still producing highly natural-sounding results. The synthesized output was rigorously evaluated through both objective metrics (Whisper-based transcription accuracy) and subjective listening tests to ensure faithful reproduction of the original voice features.

Fine-tuning CosyVoice2.0 on TIMIT corpus

To enhance the fidelity and intelligibility of re-synthesized speech while utilizing the generative capacity of large speech models for artifact mitigation, CosyVoice2.0 was fine-tuned on the sentences from the TIMIT corpus used in the training set of neural-driven dual-pathway model training. The architecture comprises two core components: (i) a “Flow” module converting tokens to acoustic features via Text Embedding, Speaker Embedding (linear layer), Encoder, and causal conditional CFM Decoder; and (ii) a “Hift” vocoder synthesizing waveforms from Flow outputs using an F0 extractor, Up-sampler, and 9 residual blocks. Selective parameter optimization was applied to: (i) the first and final layers of the 12-layer Flow decoder, and (ii) the Speaker Embedding linear layer. The model mapped speech autoencoder outputs and punctuation-free plain text to ground-truth waveforms, optimized with a compound loss consists of STFT-based KL-divergence and MFCC loss: These models were trained using AdamW optimizer with a fixed learning rate of 10-5, and set β1 = 0.9, β2 = 0.999, and ε = 10−8.

eLife Assessment

This study presents a valuable advance in reconstructing naturalistic speech from intracranial ECoG data using a dual-pathway model. The evidence supporting the claims of the authors is solid, although the rationale for employing a smaller language model rather than a large language model (LLM) should be further clarified. This work will be of interest to cognitive neuroscientists and computer scientists/engineers working on speech reconstruction from neural data.

Reviewer #1 (Public review):

Summary:

This paper introduces a dual-pathway model for reconstructing naturalistic speech from intracranial ECoG data. It integrates an acoustic pathway (LSTM + HiFi-GAN for spectral detail) and a linguistic pathway (Transformer + Parler-TTS for linguistic content). Output from the two components is later merged via CosyVoice2.0 voice cloning. Using only 20 minutes of ECoG data per participant, the model achieves high acoustic fidelity and linguistic intelligibility.

Strengths:

(1) The proposed dual-pathway framework effectively integrates the strengths of neural-to-acoustic and neural-to-text decoding and aligns well with established neurobiological models of dual-stream processing in speech and language.

(2) The integrated approach achieves robust speech reconstruction using only 20 minutes of ECoG data per subject, demonstrating the efficiency of the proposed method.

(3) The use of multiple evaluation metrics (MOS, mel-spectrogram R², WER, PER) spanning acoustic, linguistic (phoneme and word), and perceptual dimensions, together with comparisons against noise-degraded baselines, adds strong quantitative rigor to the study.

Weaknesses:

(1) It is unclear how much the acoustic pathway contributes to the final reconstruction results, based on Figures 3B-E and 4E. Including results from Baseline 2 + CosyVoice and Baseline 3 + CosyVoice could help clarify this contribution.

(2) As noted in the limitations, the reconstruction results heavily rely on pre-trained generative models. However, no comparison is provided with state-of-the-art multimodal LLMs such as Qwen3-Omni, which can process auditory and textual information simultaneously. The rationale for using separate models (Wav2Vec for speech and TTS for text) instead of a single unified generative framework should be clearly justified. In addition, the adaptor employs an LSTM architecture for speech but a Transformer for text, which may introduce confounds in the performance comparison. Is there any theoretical or empirical motivation for adopting recurrent networks for auditory processing and Transformer-based models for textual processing?

(3) The model is trained on approximately 20 minutes of data per participant, which raises concerns about potential overfitting. It would be helpful if the authors could analyze whether test sentences with higher or lower reconstruction performance include words that were also present in the training set.

(4) The phoneme confusion matrix in Figure 4A does not appear to align with human phoneme confusion patterns. For instance, /s/ and /z/ differ only in voicing, yet the model does not seem to confuse these phonemes. Does this imply that the model and the human brain operate differently at the mechanistic level?

(5) In general, is the motivation for adopting the dual-pathway model to better align with the organization of the human brain, or to achieve improved engineering performance? If the goal is primarily engineering-oriented, the authors should compare their approach with a pretrained multimodal LLM rather than relying on the dual-pathway architecture. Conversely, if the design aims to mirror human brain function, additional analysis, such as detailed comparisons of phoneme confusion matrices, should be included to demonstrate that the model exhibits brain-like performance patterns.

Reviewer #2 (Public review):

Summary:

The study by Li et al. proposes a dual-path framework that concurrently decodes acoustic and linguistic representations from ECoG recordings. By integrating advanced pre-trained AI models, the approach preserves both acoustic richness and linguistic intelligibility, and achieves a WER of 18.9% with a short (~20-minute) recording.

Overall, the study offers an advanced and promising framework for speech decoding. The method appears sound, and the results are clear and convincing. My main concerns are the need for additional control analyses and for more comparisons with existing models.

Strengths:

(1) This speech-decoding framework employs several advanced pre-trained DNN models, reaching superior performance (WER of 18.9%) with relatively short (~20-minute) neural recording.

(2) The dual-pathway design is elegant, and the study clearly demonstrates its necessity: The acoustic pathway enhances spectral fidelity while the linguistic pathway improves linguistic intelligibility.

Weaknesses:

The DNNs used were pre-trained on large corpora, including TIMIT, which is also the source of the experimental stimuli. More generally, as DNNs are powerful at generating speech, additional evidence is needed to show that decoding performance is driven by neural signals rather than by the DNNs' generative capacity.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary

This paper introduces a dual-pathway model for reconstructing naturalistic speech from intracranial ECoG data. It integrates an acoustic pathway (LSTM + HiFi-GAN for spectral detail) and a linguistic pathway (Transformer + Parler-TTS for linguistic content). Output from the two components is later merged via CosyVoice2.0 voice cloning. Using only 20 minutes of ECoG data per participant, the model achieves high acoustic fidelity and linguistic intelligibility.

Strengths

(1) The proposed dual-pathway framework effectively integrates the strengths of neural-to-acoustic and neural-to-text decoding and aligns well with established neurobiological models of dual-stream processing in speech and language.

(2) The integrated approach achieves robust speech reconstruction using only 20 minutes of ECoG data per subject, demonstrating the efficiency of the proposed method.

(3) The use of multiple evaluation metrics (MOS, mel-spectrogram R², WER, PER) spanning acoustic, linguistic (phoneme and word), and perceptual dimensions, together with comparisons against noisedegraded baselines, adds strong quantitative rigor to the study.

We thank Reviewer #1 for the supportive comments. In addition, we appreciate Reviewer #1’s thoughtful comments and feedback. By addressing these comments, we believe we have greatly improved the clarity of our claims and methodology. Below we list our point-to-point responses addressing concerns raised by Reviewer #1.

Weaknesses:

(1) It is unclear how much the acoustic pathway contributes to the final reconstruction results, based on Figures 3B-E and 4E. Including results from Baseline 2 + CosyVoice and Baseline 3 + CosyVoice could help clarify this contribution.

We sincerely appreciate the inquiry from Reviewer 1. We thank the reviewer for this suggestion. However, we believe that directly applying CosyVoice to the outputs of Baseline 2 or Baseline 3 in isolation is not methodologically feasible and would not correctly elucidate the contribution of the auditory pathway and might lead to misinterpretation.

The role of CosyVoice 2.0 in our framework is specifically voice cloning and fusion, not standalone enhancement. It is designed to integrate information from two pathways. Its operation requires two key inputs:

(1) A voice reference speech that provides the target speaker's timbre and prosodic characteristics. In our final pipeline, this is provided by the denoised output of the acoustic pathway (Baseline 2).

(2) A target word sequence that specifies the linguistic content to be spoken. This is obtained by transcribing the output of the linguistic pathway (Baseline 3) using Whisper ASR. Therefore, the standalone outputs of Baseline 2 and Baseline 3 are the purest demonstrations of what each pathway contributes before fusion. The significant improvement in WER/PER and MOS in the final output (compared to Baseline 2) and the significant improvement in melspectrogram R² (compared to Baseline 3) together demonstrate the complementary contributions of the two pathways. The fusion via CosyVoice is the mechanism that allows these contributions to be combined. We have added a clearer explanation of CosyVoice's role and the rationale for not testing it on individual baselines in the revised manuscript (Results section: "The fine-tuned voice cloner further enhances...").

Edits:

Page 11, Lines 277-282:

“ Voice cloning is used to bridge the gap between acoustic fidelity and linguistic intelligibility in speech reconstruction. This approach strategically combines the strengths of complementary pathways: the acoustic pathway preserves speaker-specific spectral characteristics while the linguistic pathway maintains lexical and phonetic precision. By integrating these components through neural voice cloning, we achieve balanced reconstruction that overcomes the limitations inherent in isolated systems. CosyVoice 2.0, the voice cloner module serves specifically as a voice cloning and fusion engine, requiring two inputs: (1) a voice reference speech (provided by the denoised output of the acoustic pathway) to specify the target speaker's identity, and (2) a target word sequence (transcribed from the output of the linguistic pathway) to specify the linguistic content. The standalone baseline outputs of the two pathways can be integrated in this way.”

(2) As noted in the limitations, the reconstruction results heavily rely on pre-trained generative models. However, no comparison is provided with state-of-the-art multimodal LLMs such as Qwen3-Omni, which can process auditory and textual information simultaneously. The rationale for using separate models (Wav2Vec for speech and TTS for text) instead of a single unified generative framework should be clearly justified. In addition, the adaptor employs an LSTM architecture for speech but a Transformer for text, which may introduce confounds in the performance comparison. Is there any theoretical or empirical motivation for adopting recurrent networks for auditory processing and Transformer-based models for textual processing?

We thank the reviewer for the insightful suggestion regarding multimodal large language models (LLMs) such as Qwen3-Omni. It is important to clarify the distinction between general-purpose interactive multimodal models and models specifically designed for high-fidelity voice cloning and speech synthesis.

As for the comparison with the state-of-the-art multimodal LLMs:

Qwen3-Omni and GLM-4-Voice are powerful conversational agents capable of processing multiple modalities including text, speech, image, and video, as described in its documentation (see: https://help.aliyun.com/zh/model-studio/qwen-tts-realtime and https://docs.bigmodel.cn/cn/guide/models/sound-and-video/glm-4-voice). However, it is primarily optimized for interactive dialogue and multimodal understanding rather than for precise, speaker-adaptive speech reconstruction from neural signals. In contrast, CosyVoice 2.0, developed by the same team at Alibaba, is specifically designed for voice cloning and text-to-speech synthesis (see: https://help.aliyun.com/zh/model-studio/text-to-speech). It incorporates advanced speaker adaptation and acoustic modeling capabilities that are essential for reconstructing naturalistic speech from limited neural data. Therefore, our choice of CosyVoice for the final synthesis stage aligns with the goal of integrating acoustic fidelity and linguistic intelligibility, which is central to our study.

For the selection of LSTM and Transformer in the two pathways:

The goal of the acoustic adaptor is to reconstruct fine-grained spectrotemporal details (formants, harmonic structures, prosodic contours) with millisecond-to-centisecond precision. These features rely heavily on local temporal dynamics and short-to-medium range dependencies (e.g., within and between phonemes/syllables). In our ablation studies (to be added in the supplementary), we found that Transformer-based adaptors, which inherently emphasize global sentence-level context through self-attention, tended to oversmooth the reconstructed acoustic features, losing critical fine-temporal details essential for naturalness. In contrast, the recurrent nature of LSTMs, with their inherent temporal state propagation, proved more effective at modeling these local sequential dependencies without excessive smoothing, leading to higher mel-spectrogram fidelity. This aligns with the neurobiological observation that early auditory cortex processes sound with precise temporal fidelity. Moreover, from an engineering perspective, LSTM-based decoders have been empirically shown to perform well in sequential prediction tasks with limited data, as evidenced in prior work on sequence modeling and neural decoding (1).

The goal of the linguistic adaptor is to decode abstract, discrete word tokens. This task benefits from modeling long-range contextual dependencies across a sentence to resolve lexical ambiguity and syntactic structure (e.g., subject-verb agreement). The self-attention mechanism of Transformers is exceptionally well-suited for capturing these global relationships, as evidenced by their dominance in NLP. Our experiments confirmed that a Transformer adaptor outperformed an LSTM-based one in word token prediction accuracy.

While a unified multimodal LLM could in principle handle both modalities, such models often face challenges in modality imbalance and task specialization. Audio and text modalities have distinct temporal scales, feature distributions, and learning dynamics. By decoupling them into separate pathways with specialized adaptors, we ensure that each modality is processed by an architecture optimized for its inherent structure. This divide-and-conquer strategy avoids the risk of one modality dominating or interfering with the learning of the other, leading to more stable training and better final performance, especially important when adapting to limited neural data.

Edits:

Page 9, Lines 214-223:

“The acoustic pathway, implemented through a bi-directional LSTM neural adaptor architecture (Fig. 1B), specializes in reconstructing fundamental acoustic properties of speech. This module directly processes neural recordings to generate precise time-frequency representations, focusing on preserving speaker-specific spectral characteristics like formant structures, harmonic patterns, and spectral envelope details. Quantitative evaluation confirms its core competency: achieving a mel-spectrogram R² of 0.793 ± 0.016 (Fig. 3B) demonstrates remarkable fidelity in reconstructing acoustic microstructure. This performance level is statistically indistinguishable from original speech degraded by 0dB additive noise (0.771 ± 0.014, p = 0.242, one-sided t-test). We chose a bidirectional LSTM architecture for this adaptor because its recurrent nature is particularly suited to modeling the fine-grained, short- to medium-range temporal dependencies (e.g., within and between phonemes and syllables) that are critical for acoustic fidelity. An ablation study comparing LSTM against Transformerbased adaptors for this task confirmed that LSTMs yielded superior mel-spectrogram reconstruction fidelity (higher R²), as detailed in Table S1, likely by avoiding the oversmoothing of spectrotemporal details sometimes induced by the strong global context modeling of Transformers”.

“To confirm that the acoustic pathway’s output is causally dependent on the neural signal rather than the generative prior of the HiFi-GAN, we performed a control analysis in which portions of the input ECoG recording were replaced with Gaussian noise. When either the first half, second half, or the entirety of the neural input was replaced by noise, the melspectrogram R² of the reconstructed speech dropped markedly, corresponding to the corrupted segment (Fig. S5). This demonstrates that the reconstruction is temporally locked to the specific neural input and that the model does not ‘hallucinate’ spectrotemporal structure from noise. These results validate that the acoustic pathway performs genuine, input-sensitive neural decoding”.

Edits:

Page 10, Lines 272-277:

“We employed a Transformer-based Seq2Seq architecture for this adaptor to effectively capture the long-range contextual dependencies across a sentence, which are essential for resolving lexical ambiguity and syntactic structure during word token decoding. This choice was validated by an ablation study (Table S2), indicating that the Transformer adaptor outperformed an LSTM-based counterpart in word prediction accuracy”

(3) The model is trained on approximately 20 minutes of data per participant, which raises concerns about potential overfitting. It would be helpful if the authors could analyze whether test sentences with higher or lower reconstruction performance include words that were also present in the training set.

Thank you for raising the important concern regarding potential overfitting given the limited size of our training dataset (~20 minutes per participant). To address this point directly, we performed a detailed lexical overlap analysis between the training and test sets.

The test set contains 219 unique words. Among these:

127 words (58.0%) appeared in the training set (primarily high-frequency, common words).

92 words (42.0%) were entirely novel and did not appear in the training set. We further examined whether trials with the best reconstruction (WER = 0) relied more on training vocabulary. Among these top-performing trials, 55.0% of words appeared in the training set. In contrast, the worst-performing trials showed 51.9% overlap in words in the training set. No significant difference was observed, suggesting that performance is not driven by simple lexical memorization.

The presence of a substantial proportion of novel words (42%) in the test set, combined with the lack of performance advantage for overlapping content, provides strong evidence that our model is generalizing linguistic and acoustic patterns rather than merely memorizing the training vocabulary. High reconstruction performance on unseen words would be improbable under severe overfitting.

Therefore, we conclude that while some lexical overlap exists (as expected in natural language), the model’s performance is driven by its ability to decode generalized neural representations, effectively mitigating the overfitting risk highlighted by the reviewer.

(4) The phoneme confusion matrix in Figure 4A does not appear to align with human phoneme confusion patterns. For instance, /s/ and /z/ differ only in voicing, yet the model does not seem to confuse these phonemes. Does this imply that the model and the human brain operate differently at the mechanistic level?

We thank the reviewer for this detailed observation regarding the difference between our model's phoneme confusion patterns and typical human perceptual confusions (e.g., the lack of /s/-/z/ confusion).

The reviewer is correct in inferring a mechanistic difference. This divergence is primarily attributable to the Parler-TTS model acting as a powerful linguistic prior. Our linguistic pathway decodes word tokens, which Parler-TTS then converts to speech. Trained on massive corpora to produce canonical pronunciations, Parler-TTS effectively performs an implicit "error correction." For instance, if the neural decoding is ambiguous between the words "sip" and "zip," the TTS model's strong prior for lexical and syntactic context will likely resolve it to the correct word, thereby suppressing purely acoustic confusions like voicing.

This has important implications for interpreting our model's errors and its relationship to brain function. The phoneme errors in our final output reflect a combination of neural decoding errors and the generative biases of the TTS model, which is optimized for intelligibility rather than mimicking raw human misperception. This does imply our model operates differently from the human auditory periphery. The human brain may first generate a percept with acoustic confusions, which higher-level language regions then disambiguate. Our model effectively bypasses the "confused percept" stage by directly leveraging a pre-trained, high-level language model for disambiguation. This is a design feature contributing to its high intelligibility, not necessarily a flaw. This observation raises a fascinating question: Could a model that more faithfully simulates the hierarchical processing of the human brain (including early acoustic confusions) provide a better fit to neural data at different processing stages? Future work could further address this question.

Edits:

add another paragraph in Discussion (Page 14, Lines 397-398):

“The phoneme confusion pattern observed in our model output (Fig. 4A) differs from classic human auditory confusion matrices. We attribute this divergence primarily to the influence of the Parler-TTS model, which serves as a strong linguistic prior in our pipeline. This component is trained to generate canonical speech from text tokens. When the upstream neural decoding produces an ambiguous or erroneous token sequence, the TTS model’s internal language model likely performs an implicit ‘error correction,’ favoring linguistically probable words and pronunciations. This underscores that our model’s errors arise from a complex interaction between neural decoding fidelity and the generative biases of the synthesis stage”

(5) In general, is the motivation for adopting the dual-pathway model to better align with the organization of the human brain, or to achieve improved engineering performance? If the goal is primarily engineeringoriented, the authors should compare their approach with a pretrained multimodal LLM rather than relying on the dual-pathway architecture. Conversely, if the design aims to mirror human brain function, additional analysis, such as detailed comparisons of phoneme confusion matrices, should be included to demonstrate that the model exhibits brain-like performance patterns.

Our primary motivation is engineering improvement, to overcome the fundamental trade-off between acoustic fidelity and linguistic intelligibility that has limited previous neural speech decoding work. The design is inspired by the related works of the convergent representation of speech and language perception (2). However, we do not claim that our LSTM and Transformer adaptors precisely simulate the specific neural computations of the human ventral and dorsal streams. The goal was to build a high-performance, data-efficient decoder. We will clarify this point in the Introduction and Discussion, stating that while the architecture is loosely inspired by previous neuroscience results, its primary validation is its engineering performance in achieving state-of-the-art reconstruction quality with minimal data.

Edits:

Page 14, Line 358-373:

“In this study, we present a dual-path framework that synergistically decodes both acoustic and linguistic speech representations from ECoG signals, followed by a fine-tuned zero-shot text-to-speech network to re-synthesize natural speech with unprecedented fidelity and intelligibility. Crucially, by integrating large pre-trained generative models into our acoustic reconstruction pipeline and applying voice cloning technology, our approach preserves acoustic richness while significantly enhancing linguistic intelligibility beyond conventional methods. Our dual-pathway architecture, while inspired by converging neuroscience insights on speech and language perception, was principally designed and validated as an engineering solution. The primary goal to build a practical decoder that achieves state-of-theart reconstruction quality with minimal data. The framework's success is therefore ultimately judged by its performance metrics, high intelligibility (WER, PER), acoustic fidelity (melspectrogram R²), and perceptual quality (MOS), which directly address the core engineering challenge we set out to solve. Using merely 20 minutes of ECoG recordings, our model achieved superior performance with a WER of 18.9% ± 3.3% and PER of 12.0% ± 2.5% (Fig. 2D, E). This integrated architecture, combining pre-trained acoustic (Wav2Vec2.0 and HiFiGAN) and linguistic (Parler-TTS) models through lightweight neural adaptors, enables efficient mapping of ECoG signals to dual latent spaces. Such methodology substantially reduces the need for extensive neural training data while achieving breakthrough word clarity under severe data constraints. The results demonstrate the feasibility of transferring the knowledge embedded in speech-data pre-trained artificial intelligence (AI) models into neural signal decoding, paving the way for more advanced brain-computer interfaces and neuroprosthetics”.

Reviewer #2 (Public review):

Summary:

The study by Li et al. proposes a dual-path framework that concurrently decodes acoustic and linguistic representations from ECoG recordings. By integrating advanced pre-trained AI models, the approach preserves both acoustic richness and linguistic intelligibility, and achieves a WER of 18.9% with a short (~20-minute) recording.

Overall, the study offers an advanced and promising framework for speech decoding. The method appears sound, and the results are clear and convincing. My main concerns are the need for additional control analyses and for more comparisons with existing models.

Strengths:

(1) This speech-decoding framework employs several advanced pre-trained DNN models, reaching superior performance (WER of 18.9%) with relatively short (~20-minute) neural recording.

(2) The dual-pathway design is elegant, and the study clearly demonstrates its necessity: The acoustic pathway enhances spectral fidelity while the linguistic pathway improves linguistic intelligibility.

We thank Reviewer #2 for supportive comments. In addition, we appreciate Reviewer #2’s thoughtful comments and feedback. By addressing these comments, we believe we have greatly improved the clarity of our claims and methodology. Below we list our point-to-point responses addressing concerns raised by Reviewer #2.

Weaknesses:

The DNNs used were pre-trained on large corpora, including TIMIT, which is also the source of the experimental stimuli. More generally, as DNNs are powerful at generating speech, additional evidence is needed to show that decoding performance is driven by neural signals rather than by the DNNs' generative capacity.

Thank you for raising this crucial point regarding the potential for pre-trained DNNs to generate speech independently of the neural input. We fully agree that it is essential to disentangle the contribution of the neural signals from the generative priors of the models. To address this directly, we have conducted two targeted control analyses, as you suggested, and have integrated the results into the revised manuscript (see Fig. S5 and the corresponding description in the Results section):

(1) Random noise input: We fed Gaussian noise (matched in dimensionality and temporal structure to real ECoG recordings) into the trained adaptors. The outputs were acoustically unstructured and linguistically incoherent, confirming that the generative models alone cannot produce meaningful speech without valid neural input.

(2) Partial sentence input (real + noise): For the acoustic pathway, we systematically replaced portions of the ECoG input with noise. The reconstruction quality (mel-spectrogram R²) dropped significantly in the corrupted segments, demonstrating that the decoding is temporally locked to the neural signal and does not “hallucinate” speech from noise.

These results provide strong evidence that our model’s performance is causally dependent on and sensitive to the specific neural input, validating that it performs genuine neural decoding rather than merely leveraging the generative capacity of the pre-trained DNNs.

The detailed edits are in the “recommendations” below. (See recommendations (1) and (2))

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Clarify the results shown in Figure 4E. The integrated approach appears to perform comparably to Baseline 3 in phoneme class clarity. However, Baseline 3 represents the output of the linguistic pathway alone, which is expected to encode information primarily at the word level.

We appreciate the reviewer's observation and agree that clarification is needed. The phoneme class clarity (PCC) metric shown in Figure 4E measures whether mis-decoded phonemes are more likely to be confused within their own class (vowel-vowel or consonantconsonant) rather than across classes (vowel-consonant). A higher PCC indicates that the model's errors tend to be phonologically similar sounds (e.g., one vowel mistaken for another), which is a reasonable property for intelligibility.

We would like to clarify the nature of Baseline 3. As stated in the manuscript (Results section: "The linguistic pathway reconstructs high-intelligibility, higher-level linguistic information"), Baseline 3 is the output of our linguistic pathway. This pathway operates as follows: the ECoG signals are mapped to word tokens via the Transformer adaptor, and these tokens are then synthesized into speech by the frozen Parler-TTS model. Crucially, the input to Parler-TTS is a sequence of word tokens.

It is important to distinguish between the levels of performance measured: Word Error Rate (WER) reflects accuracy at the lexical level (whole words). The linguistic pathway achieves a low WER by design, as it directly decodes word sequences. Phoneme Error Rate (PER) reflects accuracy at the sublexical phonetic level (phonemes). A low WER generally implies a low PER, because robust word recognition requires reliable phoneme-level representations within the TTS model's prior. This explains why Baseline 3 also exhibits a low PER. However, acoustic fidelity (captured by metrics like mel-spectrogram R²) requires the preservation of fine-grained spectrotemporal details such as pitch, timbre, prosody, and formant structures, information that is not directly encoded at the lexical level and is therefore not a strength of the purely linguistic pathway.

While Parler-TTS internally models sub-word/phonetic information to generate the acoustic waveform, the primary linguistic information driving the synthesis is at the lexical (word) level. The generated speech from Baseline 3 therefore contains reconstructed phonemic sequences derived from the decoded word tokens, not from direct phoneme-level decoding of ECoG.

Therefore, the comparable PCC between our final integrated model and Baseline 3 (linguistic pathway) suggests that the phoneme-level error patterns (i.e., the tendency to confuse within-class phonemes) in our final output are largely inherited from the high-quality linguistic prior embedded in the pre-trained TTS model (Parler-TTS). The integrated framework successfully preserves this desirable property from the linguistic pathway while augmenting it with speaker-specific acoustic details from the acoustic pathway, thereby achieving both high intelligibility (low WER/PER) and high acoustic fidelity (high melspectrogram R²).

We will revise the caption of Figure 4E and the corresponding text in the Results section to make this interpretation explicit.

Edits:

Page 12, Lines 317-322:

“In addition to the confusion matrices, we categorized the phonemes into vowels and consonants to assess the phoneme class clarity. We defined "phoneme class clarity" (PCC) as the proportion of errors where a phoneme was misclassified within the same class versus being misclassified into a different class. The purpose of introducing PCC is to demonstrate that most of the misidentified phonemes belong to the same category (confusion between vowels or consonants), rather than directly comparing the absolute accuracy of phoneme recognition. For instance, a vowel being mistaken for another vowel would be considered a within-class error, whereas a vowel being mistaken for a consonant would be classified as a between-class error”

(2) Add results from Baseline 2 + CosyVoice and Baseline 3 + CosyVoice to clarify the contribution of the auditory pathway.

Thank you for the suggestion. We appreciate the opportunity to clarify the role of CosyVoice in our framework.

As explained in our response to point (1), CosyVoice 2.0 is designed as a fusion module that requires two inputs: 1) a voice reference (from the acoustic pathway) to specify speaker identity, and 2) a word sequence (from the linguistic pathway) to specify linguistic content. Because it is not a standalone enhancer, applying CosyVoice to a single pathway output (e.g., Baseline 2 or 3 alone) is not quite feasible and would not reflect its intended function and could lead to misinterpretation of each pathway’s contribution.

Instead, we have evaluated the contribution of each pathway by comparing the final integrated output against each standalone pathway output (Baseline 2 and 3). The significant improvements in both acoustic fidelity and linguistic intelligibility demonstrate the complementary roles of the two pathways, which are effectively fused through CosyVoice.

(3) Justify your choice of using LSTM and Transformer architecture for the auditory and linguistic neural adaptors, respectively, and how your methods could compare to using a unified generative multimodal LLM for both pathways.

Thank you for revisiting this important point. We appreciate your interest in the architectural choices and their relationship to state-of-the-art multimodal models.

As detailed in our response to point (2), our choice of LSTM for the acoustic pathway and Transformer for the linguistic pathway is driven by task-specific requirements, supported by ablation studies (Supplementary Tables 1–2). The acoustic pathway benefits from LSTM’s ability to model fine-grained, local temporal dependencies without over-smoothing. The linguistic pathway benefits from Transformer’s ability to capture long-range semantic and syntactic context.

Regarding comparison with unified multimodal LLMs (e.g., Qwen3-Omni), we clarified that such models are optimized for interactive dialogue and multimodal understanding, while our framework relies on specialist models (CosyVoice 2.0, Parler-TTS) that are explicitly designed for high-fidelity, speaker-adaptive speech synthesis, a requirement central to our decoding task.

We have incorporated these justifications into the revised manuscript (Results and Discussion sections) and appreciate the opportunity to further emphasize these points.

Edits:

Page 9, Lines 214-223:

“The acoustic pathway, implemented through a bi-directional LSTM neural adaptor architecture (Fig. 1B), specializes in reconstructing fundamental acoustic properties of speech. This module directly processes neural recordings to generate precise time-frequency representations, focusing on preserving speaker-specific spectral characteristics like formant structures, harmonic patterns, and spectral envelope details. Quantitative evaluation confirms its core competency: achieving a mel-spectrogram R² of 0.793 ± 0.016 (Fig. 3B) demonstrates remarkable fidelity in reconstructing acoustic microstructure. This performance level is statistically indistinguishable from original speech degraded by 0dB additive noise (0.771 ± 0.014, p = 0.242, one-sided t-test). We chose a bidirectional LSTM architecture for this adaptor because its recurrent nature is particularly suited to modeling the fine-grained, short- to medium-range temporal dependencies (e.g., within and between phonemes and syllables) that are critical for acoustic fidelity. An ablation study comparing LSTM against Transformerbased adaptors for this task confirmed that LSTMs yielded superior mel-spectrogram reconstruction fidelity (higher R²), as detailed in Table S1, likely by avoiding the oversmoothing of spectrotemporal details sometimes induced by the strong global context modeling of Transformers”.

“To confirm that the acoustic pathway’s output is causally dependent on the neural signal rather than the generative prior of the HiFi-GAN, we performed a control analysis in which portions of the input ECoG recording were replaced with Gaussian noise. When either the first half, second half, or the entirety of the neural input was replaced by noise, the melspectrogram R² of the reconstructed speech dropped markedly, corresponding to the corrupted segment (Fig. S5). This demonstrates that the reconstruction is temporally locked to the specific neural input and that the model does not ‘hallucinate’ spectrotemporal structure from noise. These results validate that the acoustic pathway performs genuine, input-sensitive neural decoding”.

Page 10, Lines 272-277:

“We employed a Transformer-based Seq2Seq architecture for this adaptor to effectively capture the long-range contextual dependencies across a sentence, which are essential for resolving lexical ambiguity and syntactic structure during word token decoding. This choice was validated by an ablation study (Table S2), indicating that the Transformer adaptor outperformed an LSTM-based counterpart in word prediction accuracy”.

(4) Discuss the differences between the model's phoneme confusion matrix in Figure 4A and human phoneme confusion patterns. In addition, please clarify whether the adoption of the dual-pathway architecture is primarily intended to simulate the organization of the human brain or to achieve engineering improvements.

The observed difference between our model's phoneme confusion matrix and typical human perceptual confusion patterns (e.g., the noted lack of confusion between /s/ and /z/) is, as the reviewer astutely infers, likely attributable to the TTS model (Parler-TTS) acting as a powerful linguistic prior. The linguistic pathway decodes word tokens, and Parler-TTS converts these tokens into speech. Parler-TTS is trained on massive text and speech corpora to produce canonical, clean pronunciations. It effectively performs a form of "error correction" or "canonicalization" based on its internal language model. For example, if the neural decoding is ambiguous between "sip" and "zip", the TTS model's strong prior for lexical and syntactic context may robustly resolve it to the correct word, suppressing purely acoustic confusions like voicing. Therefore, the phoneme errors in our final output reflect a combination of neural decoding errors and the TTS model's generation biases, which are optimized for intelligibility rather than mimicking human misperception. We will add this explanation to the paragraph discussing Figure 4A.

Our primary motivation is engineering improvement, to overcome the fundamental tradeoff between acoustic fidelity and linguistic intelligibility that has limited previous neural speech decoding work. The design is inspired by the convergent representation of speech and language perception (1). However, we do not claim that our LSTM and Transformer adaptors precisely simulate the specific neural computations of the human ventral and dorsal streams. The goal was to build a high-performance, data-efficient decoder. We will clarify this point in the Introduction and Discussion, stating that while the architecture is loosely inspired by previous neuroscience results, its primary validation is its engineering performance in achieving state-of-the-art reconstruction quality with minimal data.

Edits:

Pages 2-3, Lines 74-85:

“Here, we propose a unified and efficient dual-pathway decoding framework that integrates the complementary strengths of both paradigms to enhance the performance of re-synthesized natural speech from the engineering performance. Our method maps intracranial electrocorticography (ECoG) signals into the latent spaces of pre-trained speech and language models via two lightweight neural adaptors: an acoustic pathway, which captures low-level spectral features for naturalistic speech synthesis, and a linguistic pathway, which extracts high-level linguistic tokens for semantic intelligibility. These pathways are fused using a finetuned text-to-speech (TTS) generator with voice cloning, producing re-synthesized speech that retains both the acoustic spectrotemporal details, such as the speaker’s timbre and prosody, and the message linguistic content. The adaptors rely on near-linear mappings and require only 20 minutes of neural data per participant for training, while the generative modules are pre-trained on large unlabeled corpora and require no neural supervision”.

Page 14, Lines 358-373:

“In this study, we present a dual-path framework that synergistically decodes both acoustic and linguistic speech representations from ECoG signals, followed by a fine-tuned zero-shot text-to-speech network to re-synthesize natural speech with unprecedented fidelity and intelligibility. Crucially, by integrating large pre-trained generative models into our acoustic reconstruction pipeline and applying voice cloning technology, our approach preserves acoustic richness while significantly enhancing linguistic intelligibility beyond conventional methods. Our dual-pathway architecture, while inspired by converging neuroscience insights on speech and language perception, was principally designed and validated as an engineering solution. The primary goal to build a practical decoder that achieves state-of-the-art reconstruction quality with minimal data. The framework's success is therefore ultimately judged by its performance metrics, high intelligibility (WER, PER), acoustic fidelity (mel-spectrogram R²), and perceptual quality (MOS), which directly address the core engineering challenge we set out to solve. Using merely 20 minutes of ECoG recordings, our model achieved superior performance with a WER of 18.9% ± 3.3% and PER of 12.0% ± 2.5% (Fig. 2D, E). This integrated architecture, combining pre-trained acoustic (Wav2Vec2.0 and HiFi-GAN) and linguistic (Parler-TTS) models through lightweight neural adaptors, enables efficient mapping of ECoG signals to dual latent spaces. Such methodology substantially reduces the need for extensive neural training data while achieving breakthrough word clarity under severe data constraints. The results demonstrate the feasibility of transferring the knowledge embedded in speech-data pre-trained artificial intelligence (AI) models into neural signal decoding, paving the way for more advanced brain-computer interfaces and neuroprosthetics”.

Reviewer #2 (Recommendations for the authors):

(1) My main question is whether any experimental stimuli overlap with the data used to pre-train the models. The authors might consider using pre-trained models trained on other corpora and training their own model without the TIMIT corpus. Additionally, as pretrained models were used, it might be helpful to evaluate to what extent the decoding is sensitive to the input neural recording or whether the model always outputs meaningful speech. The authors might consider two control analyses: a) whether the model still generates speech-like output if the input is random noise; b) whether the model can decode a complete sentence if the first half recording of a sentence is real but the second half is replaced with noise.

We thank the reviewer for raising this crucial point regarding potential data leakage and the sensitivity of decoding to neural input.

We confirm that the pre-training phase of our core models (Wav2Vec2.0 encoder, HiFiGAN decoder) was conducted exclusively on the LibriSpeech corpus (960 hours), which is entirely separate from the TIMIT corpus used for our ECoG experiments. The subsequent fine-tuning of the CosyVoice 2.0 voice cloner for speaker adaptation was performed on the training set portion of the entire TIMIT corpus. Importantly, the test set for all neural decoding evaluations was strictly held out and never used during any fine-tuning stage. This data separation is now explicitly stated in the " Methods" sections for the Speech Autoencoder and the CosyVoice fine-tuning.

Regarding the potential of training on other corpora, we agree it is a valuable robustness check. Previous work has demonstrated that self-supervised speech models like Wav2Vec2.0 learn generalizable representations that transfer well across domains (e.g., Millet et al., NeurIPS 2022). We believe our use of LibriSpeech, a large and diverse corpus, provides a strong, general-purpose acoustic prior.

We agree with the reviewer that control analyses are essential to demonstrate that the decoded output is driven by neural signals and not merely the generative prior of the models. We have conducted the following analyses and will include them in the revised manuscript (likely in a new Supplementary Figure or Results subsection):

(a) Random Noise Input: We fed Gaussian noise (matched in dimensionality and temporal length to the real ECoG input) into the trained acoustic and linguistic adaptors. The outputs were evaluated. The acoustic pathway generated unstructured, noisy spectrograms with no discernible phonetic structure, and the linguistic pathway produced either highly incoherent word sequences or failed to generate meaningful tokens. The fusion via CosyVoice produced unintelligible babble. This confirms that the generative models alone cannot produce structured speech without meaningful neural input.

(b) Partial Sentence Input (Real + Noise): In the acoustic pathway, we replaced the first half, the second half, and all the ECoG recording for test sentences with Gaussian noise. The melspectrogram R2 showed a clear degradation in the reconstructed speech corresponding to the noisy segment. We did not do similar experiments in the linguistic pathway because the TTS generator is pre-trained by HuggingFace. We did not train any parameters of Parler-TTS. These results strongly indicate that our model's performance is contingent on and sensitive to the specific neural input, validating that it is performing genuine neural decoding.

Edits:

Page 19, Lines 533-538:

“The parameters in Wav2Vec2.0 were frozen within this training phase. The parameters in HiFi-GAN were optimized using the Adam optimizer with a fixed learning rate of 10-5, 𝛽! = 0.9, 𝛽2 = 0.999. We trained this Autoencoder in LibriSpeech, a 960-hour English speech corpus with a sampling rate of 16kHz, which is entirely separate from the TIMIT corpus used for our ECoG experiments. We spent 12 days in parallel training on 6 Nvidia GeForce RTX3090 GPUs. The maximum training epoch was 2000. The optimization did not stop until the validation loss no longer decreased”.

Edits:

Page9, Lines214-223:

“The acoustic pathway, implemented through a bi-directional LSTM neural adaptor architecture (Fig. 1B), specializes in reconstructing fundamental acoustic properties of speech. This module directly processes neural recordings to generate precise time-frequency representations, focusing on preserving speaker-specific spectral characteristics like formant structures, harmonic patterns, and spectral envelope details. Quantitative evaluation confirms its core competency: achieving a mel-spectrogram R² of 0.793 ± 0.016 (Fig. 3B) demonstrates remarkable fidelity in reconstructing acoustic microstructure. This performance level is statistically indistinguishable from original speech degraded by 0dB additive noise (0.771 ± 0.014, p = 0.242, one-sided t-test). We chose a bidirectional LSTM architecture for this adaptor because its recurrent nature is particularly suited to modeling the fine-grained, short- to medium-range temporal dependencies (e.g., within and between phonemes and syllables) that are critical for acoustic fidelity. An ablation study comparing LSTM against Transformer-based adaptors for this task confirmed that LSTMs yielded superior mel-spectrogram reconstruction fidelity (higher R²), as detailed in Table S1, likely by avoiding the oversmoothing of spectrotemporal details sometimes induced by the strong global context modeling of Transformers”.

“To confirm that the acoustic pathway’s output is causally dependent on the neural signal rather than the generative prior of the HiFi-GAN, we performed a control analysis in which portions of the input ECoG recording were replaced with Gaussian noise. When either the first half, second half, or the entirety of the neural input was replaced by noise, the melspectrogram R² of the reconstructed speech dropped markedly, corresponding to the corrupted segment (Fig. S5). This demonstrates that the reconstruction is temporally locked to the specific neural input and that the model does not ‘hallucinate’ spectrotemporal structure from noise. These results validate that the acoustic pathway performs genuine, input-sensitive neural decoding”

(2) For BCI applications, the decoding speed matters. Please report the model's inference speed. Additionally, the authors might also consider reporting cross-participant generalization and how the accuracy changes with recording duration.

We thank the reviewer for these practical and important suggestions.

Inference Speed: You are absolutely right. On our hardware (single NVIDIA GeForce RTX 3090 GPU), the current pipeline has an inference time that is longer than the duration of the target speech segment. The primary bottlenecks are the sequential processing in the autoregressive linguistic adaptor and the high-resolution waveform generation in CosyVoice 2.0. This latency currently limits real-time application. We have now added this in the Discussion acknowledging this limitation and stating that future work must focus on architectural optimizations (e.g., non-autoregressive models, lighter vocoders) and potential hardware acceleration to achieve real-time performance, which is critical for a practical BCI.

Cross-Participant Generalization: We agree that this is a key question for scalability. Our framework already addresses part of the cross-participant generalization challenge through the use of pre-trained generative modules (HiFi-GAN, Parler-TTS, CosyVoice 2.0), which are pretrained on large corpora and shared across all participants. Only a small fraction of the model, the lightweight neural adaptors, is subject-specific and requires a small amount of supervised fine-tuning (~20 minutes per participant). This design significantly reduces the per-subject calibration burden. As the reviewer implies, the ultimate goal would be pure zero-shot generalization. A promising future direction is to further improve cross-participant alignment by learning a shared neural feature encoder (e.g., using contrastive or self-supervised learning on aggregated ECoG data) before the personalized adaptors. We have added a paragraph in the Discussion outlining this as a major next step to enhance the framework’s practicality and further reduce calibration time.

Accuracy vs. Recording Duration: Thank you for this insightful suggestion. To systematically evaluate the impact of training data volume on performance, we have conducted additional experiments using progressively smaller subsets of the full training set (i.e., 25%, 50%, and 75%). When we used more than 50% of the training data, performance degrades gracefully rather than catastrophically with less data, which is promising for potential clinical scenarios where data collection may be limited. We add another figure (Fig. S4) to demonstrate this.

Edits:

Pages 15-16, Lines 427-452:

“There are several limitations in our study. The quality of the re-synthesized speech heavily relies on the performance of the generative model, indicating that future work should focus on refining and enhancing these models. Currently, our study utilized English speech sentences as input stimuli, and the performance of the system in other languages remains to be evaluated. Regarding signal modality and experimental methods, the clinical setting restricts us to collecting data during brief periods of awake neurosurgeries, which limits the amount of usable neural activity recordings. Overcoming this time constraint could facilitate the acquisition of larger datasets, thereby contributing to the re-synthesis of higher-quality natural speech. Furthermore, the inference speed of the current pipeline presents a challenge for real-time applications. On our hardware (a single NVIDIA GeForce RTX 3090 GPU), synthesizing speech from neural data takes approximately two to three times longer than the duration of the target speech segment itself. This latency is primarily attributed to the sequential processing in the autoregressive linguistic adaptor and the computationally intensive high-fidelity waveform generation in the vocoder (CosyVoice 2.0). While the current study focuses on offline reconstruction accuracy, achieving real-time or faster-than-real-time inference is a critical engineering goal for viable speech BCI prosthetics. Future work must therefore prioritize architectural optimizations, such as exploring non-autoregressive decoding strategies and more efficient neural vocoders, alongside potential hardware acceleration. Additionally, exploring non-invasive methods represents another frontier; with the accumulation of more data and the development of more powerful generative models, it may become feasible to achieve effective non-invasive neural decoding for speech resynthesis. Moreover, while our framework adopts specialized architectures (LSTM and Transformer) for distinct decoding tasks, an alternative approach is to employ a unified multimodal large language model (LLM) capable of joint acoustic-linguistic processing. Finally, the current framework requires training participant-specific adaptors, which limits its immediate applicability for new users. A critical next step is to develop methods that learn a shared, cross-participant neural feature encoder, for instance, by applying contrastive or selfsupervised learning techniques to larger aggregated ECoG datasets. Such an encoder could extract subject-invariant neural representations of speech, serving as a robust initialization before lightweight, personalized fine-tuning. This approach would dramatically reduce the amount of per-subject calibration data and time required, enhancing the practicality and scalability of the decoding framework for real-world BCI applications”

“In summary, our dual-path framework achieves high speech reconstruction quality by strategically integrating language models for lexical precision and voice cloning for vocal identity preservation, yielding a 37.4% improvement in MOS scores over conventional methods. This approach enables high-fidelity, sentence-level speech synthesis directly from cortical recordings while maintaining speaker-specific vocal characteristics. Despite current constraints in generative model dependency and intraoperative data collection, our work establishes a new foundation for neural decoding development. Future efforts should prioritize: (1) refining few-shot adaptation techniques, (2) developing non-invasive implementations, (3) expanding to dynamic dialogue contexts, and (4) cross-subject applications. The convergence of neurophysiological data with multimodal foundation models promises transformative advances, not only revolutionizing speech BCIs but potentially extending to cognitive prosthetics for memory augmentation and emotional communication. Ultimately, this paradigm will deepen our understanding of neural speech processing while creating clinically viable communication solutions for those with severe speech impairments”

Edits:

add another section in Methods: Page 22, Line 681:

“Ablation study on training data volume”.

“To assess the impact of training data quantity on decoding performance, we conducted an additional ablation experiment. For each participant, we created subsets of the full training set corresponding to 25%, 50%, and 75% of the original data by random sampling while preserving the temporal continuity of speech segments. Personalized acoustic and linguistic adaptors were then independently trained from scratch on each subset, following the identical architecture and optimization procedures described above. All other components of the pipeline, including the frozen pre-trained generators (HiFi-GAN, Parler-TTS) and the CosyVoice 2.0 voice cloner, remained unchanged. Performance metrics (mel-spectrogram R², WER, PER) were evaluated on the same held-out test set for all data conditions. The results (Fig. S4) demonstrate that when more than 50% of the training data is utilized, performance degrades gracefully rather than catastrophically, which is a promising indicator for clinical applications with limited data collection time”.

(3) I appreciate that the author compared their model with the MLP, but more comparisons with previous models could be beneficial. Even simply summarizing some measures of earlier models, such as neural recording duration, WER, PER, etc., is ok.

Thank you for this suggestion. We agree that a broader comparison contextualizes our contribution. We also acknowledge that given the differences in tasks, signal modality, and amount of data, it’s hard to draw a direct comparison. The main goal of this table is to summarize major studies, their methods and results for reference. We have now added a new Supplementary Table that summarizes key metrics from several recent and relevant studies in neural speech decoding. The table includes:

- Neural modality (e.g., ECoG, sEEG, Utah array)

- Approximate amount of neural data used per subject for decoder training

- Primary task (perception vs. production)

-Decoding framework

-Reported Word Error Rate (WER) or similar intelligibility metrics (e.g., Character Error Rate)

-Reported acoustic fidelity metrics (if available, e.g., spectral correlation)

This table includes works such as Anumanchipalli et al., Nature 2019; Akbari et al., Sci Rep 2019; Willett et al., Nature 2023; and other contemporary studies. The table clearly shows that our dual-path framework achieves a highly competitive WER (~18.9%) using an exceptionally short neural recording duration (~20 minutes), highlighting its data efficiency. We will refer to this table in the revised manuscript.

Edits:

Page 14, Lines 374-376:

“Our framework establishes a framework for speech decoding by outperforming prior acousticonly or linguistic-only approaches (Table S3) through integrated pretraining-powered acoustic and linguistic decoding”

Minor:

(1) Some processes might be described earlier, for example, the electrodes were selected, and the model was trained separately for each participant. That information was only described in the Method section now.

Thank you for catching these. We have revised the manuscript accordingly.

Edits:

Page4, Lines 89-95:

“Our proposed framework for reconstructing speech from intracranial neural recordings is designed around two complementary decoding pathways: an acoustic pathway focused on preserving low-level spectral and prosodic detail, and a linguistic pathway focused on decoding high-level textual and semantic content. For every participant, our adaptor is independently trained, and we select speech-responsive electrodes (selection details are provided in the Methods section) to tailor the model to individual neural patterns. These two streams are ultimately fused to synthesize speech that is both natural-sounding and intelligible, capturing the full richness of spoken language. Fig. 1 provides a schematic overview of this dual-pathway architecture”

(2) Line 224-228 Figure 2 should be Figure 3

Thank you for catching these. We have revised the manuscript accordingly. The information about participant-specific training and electrode selection is now briefly mentioned in the "Results" overview (section: "The acoustic and linguistic performance..."), with details still in the Methods. The figure reference error has been corrected.

Edits:

Page7, Lines 224-228:

“However, exclusive reliance on acoustic reconstruction reveals fundamental limitations. Despite excellent spectral fidelity, the pathway produces critically impaired linguistic intelligibility. At the word level, intelligibility remains unacceptably low (WER = 74.6 ± 5.5%, Fig. 3D), while MOS and phoneme-level precision fares only marginally better (MOS = 2.878 ± 0.205, Fig. 3C; PER = 28.1 ± 2.2%, Fig. 3E)”.

(3) For Figure 3C, why does the MOS seem to be higher for baseline 3 than for ground truth? Is this significant?

This is a detailed observation. Baseline 3 achieves a mean opinion score of 4.822 ± 0.086 (Fig. 3C), significantly surpassing even the original human speech (4.234 ± 0.097, p = 6.674×10⁻33). We believe this trend arises because the TIMIT corpus, recorded decades ago, contains inherent acoustic noise and relatively lower fidelity compared to modern speech corpus. In contrast, the Parler-TTS model used in Baseline 3 is trained on massive, highquality, clean speech datasets. Therefore, it synthesizes speech that listeners may subjectively perceive as "cleaner" or more pleasant, even if it lacks the original speaker's voice. Crucially, as the reviewer implies, our final integrated output does not aim to maximize MOS at the cost of speaker identity; it successfully balances this subjective quality with high intelligibility and restored acoustic fidelity. We will add a brief note explaining this possible reason in the caption of Figure 3C.

Edits:

Page9, Lines 235-245:

“The linguistic pathway reconstructs high-intelligibility, higher-level linguistic information”

“The linguistic pathway, instantiated through a pre-trained TTS generator (Fig. 1B), excels in reconstructing abstract linguistic representations. This module operates at the phonological and lexical levels, converting discrete word tokens into continuous speech signals while preserving prosodic contours, syllable boundaries, and phonetic sequences. It achieves a mean opinion score of 4.822 ± 0.086 (Fig. 3C) - significantly surpassing even the original human speech (4.234 ± 0.097, p = 6.674×10⁻33) in that the TIMIT corpus, recorded decades ago, contains inherent acoustic noise and relatively lower fidelity compared to modern speech corpus. Complementing this perceptual quality, objective intelligibility metrics confirm outstanding performance: WER reaches 17.7 ± 3.2%, with PER at 11.0 ± 2.3%”.

Reference

(1) Chen M X, Firat O, Bapna A, et al. The best of both worlds: Combining recent advances in neural machine translation[C]//Proceedings of the 56th annual meeting of the association for computational linguistics (Volume 1: Long papers). 2018: 76-86

(2) P. Chen et al. Do Self-Supervised Speech and Language Models Extract Similar Representations as Human Brain? 2024 IEEE International Conference on Acoustics, Speech and Signal Processing (ICASSP 2024). 2225–2229 (2024).

(3) H. Akbari, B. Khalighinejad, J. L. Herrero, A. D. Mehta, N. Mesgarani, Towards reconstructing intelligible speech from the human auditory cortex. Scientific reports 9, 874 (2019).

(4) S. Komeiji et al., Transformer-Based Estimation of Spoken Sentences Using Electrocorticography. Int Conf Acoust Spee, 1311-1315 (2022).

(5) L. Bellier et al., Music can be reconstructed from human auditory cortex activity using nonlinear decoding models. Plos Biology 21, (2023).

(6) F. R. Willett et al., A high-performance speech neuroprosthesis. Nature 620, (2023).

(7) S. L. Metzger et al., A high-performance neuroprosthesis for speech decoding and avatar control. Nature 620, 1037-1046 (2023).

(8) J. W. Li et al., Neural2speech: A Transfer Learning Framework for NeuralDriven Speech Reconstruction. Int Conf Acoust Spee, 2200-2204 (2024).

(9) X. P. Chen et al., A neural speech decoding framework leveraging deep learning and speech synthesis. Nat Mach Intell 6, (2024).

(10) M. Wairagkar et al., An instantaneous voice-synthesis neuroprosthesis. Nature, (2025).

Version 2

Abstract

Reconstructing speech from neural recordings is crucial for understanding speech coding and developing brain-computer interfaces (BCIs). However, existing methods trade off acoustic richness (pitch, prosody) for linguistic intelligibility (words, phonemes). To overcome this limitation, we propose a dual-path framework to concurrently decode acoustic and linguistic representations. The acoustic pathway uses a long-short term memory (LSTM) decoder and a high-fidelity generative adversarial network (HiFi-GAN) to reconstruct spectrotemporal features. The linguistic pathway employs a transformer adaptor and text-to-speech (TTS) generator for word tokens. These two pathways merge via voice cloning to combine both acoustic and linguistic validity. Using only 20 minutes of electrocorticography (ECoG) data per subject, our approach achieves highly intelligible synthesized speech (mean opinion score = 4.0/5.0, word error rate = 18.9%). Our dual-path framework reconstructs natural and intelligible speech from ECoG, resolving the acoustic-linguistic trade-off.

Introduction

Understanding how the human brain encodes and decodes spoken language is a central challenge in cognitive neuroscience (1–5), with profound implications for brain-computer interfaces (BCIs) (6–9). Reconstructing perceived speech directly from neural activity provides a powerful approach to studying the hierarchical organization of language processing in the auditory cortex and to developing assistive technologies for individuals with impaired communication abilities (10–15). This neural decoding paradigm not only helps reveal the representational structure of linguistic features across the cortex but also enables the synthesis of intelligible speech directly from brain signals (16–18).

Recent advances in deep learning have accelerated progress in neural speech reconstruction by enabling end-to-end mappings from neural recordings to acoustic waveforms (6, 7, 10, 14, 19, 20). However, these models typically require large-scale paired datasets, hours or even days of simultaneous neural and speech recordings per subject, to learn the complex nonlinear transformations involved. This data scarcity remains a major bottleneck, especially in clinical or intracranial recording settings where data acquisition is constrained by practical and ethical limitations.

To address this challenge, recent work has leveraged pre-trained self-supervised learning (SSL) models, such as Wav2Vec2.0 and HuBERT, which learn rich acoustic and phonetic representations from large-scale unlabeled speech corpora (21, 22). These models have been shown to exhibit strong correspondence with neural activity in the auditory cortex, supporting near-linear mappings between latent model representations and cortical responses (23–25). Similarly, large language models trained purely on text, such as GPT (26), also capture high-level semantic representations that correlate with activity in language-responsive cortical regions (27–29). Together, these findings suggest that the latent spaces of speech and language models provide a natural bridge between brain activity and interpretable representations of speech, enabling more data-efficient neural decoding.

Current approaches to speech decoding from brain activity can be broadly categorized into two paradigms. The first, neural-to-acoustic decoding, maps brain signals to low-level spectral or waveform features such as formants, pitch, or spectrograms (10, 11, 15, 30). This approach captures fine acoustic detail, including speaker identity, prosody, and timbre, and benefits from pre-trained speech synthesizers for waveform generation (17–19, 31, 32). However, due to the high dimensionality and dynamic nature of speech, this regression problem is inherently difficult and often yields low intelligibility when training data is limited. The second, neural-to-text decoding, treats speech reconstruction as a temporal classification task, mapping neural signals to a sequence of discrete linguistic units such as phonemes, characters, or words (12, 14, 33–35). This strategy achieves strong performance with relatively small datasets and aligns well with natural language processing (NLP) pipelines (36–39), but sacrifices naturalness and expressiveness and lacks paralinguistic features crucial for human communication.

Here, we propose a unified and efficient dual-pathway decoding framework that integrates the complementary strengths of both paradigms to enhance the performance of re-synthesized natural speech from the engineering performance. Our method maps intracranial electrocorticography (ECoG) signals into the latent spaces of pre-trained speech and language models via two lightweight neural adaptors: an acoustic pathway, which captures low-level spectral features for naturalistic speech synthesis, and a linguistic pathway, which extracts high-level linguistic tokens for semantic intelligibility. These pathways are fused using a fine-tuned text-to-speech (TTS) generator with voice cloning, producing re-synthesized speech that retains both the acoustic spectrotemporal details, such as the speaker’s timbre and prosody, and the message linguistic content. The adaptors rely on near-linear mappings and require only 20 minutes of neural data per participant for training, while the generative modules are pre-trained on large unlabeled corpora and require no neural supervision.

Results

Overview of the proposed speech re-synthesis method

Our proposed framework for reconstructing speech from intracranial neural recordings is designed around two complementary decoding pathways: an acoustic pathway focused on preserving low-level spectral and prosodic detail, and a linguistic pathway focused on decoding high-level textual and semantic content. For every participant, our adaptor is independently trained, and we select speech-responsive electrodes (selection details are provided in the Methods section) to tailor the model to individual neural patterns. These two streams are ultimately fused to synthesize speech that is both natural-sounding and intelligible, capturing the full richness of spoken language. Fig. 1 provides a schematic overview of this dual-pathway architecture.

The neural data acquisition and acoustic-linguistic dual-pathway framework for neural-driven natural speech re-synthesis.

(A) The neural data acquisition. We collected ECoG data from 9 monolingual native English participants as they each listened to English sentences from the TIMIT corpus, resulting in 20 minutes of neural recording data per participant. Furthermore, for each participant, 70%, 20%, and 10% of the recorded neural activities are respectively allocated to the training, validation, and test sets randomly. (B) The acoustic-linguistic dual-pathway framework. The acoustic pathway consists of two stages: Stage 1: A HiFi-GAN generator is pre-trained to synthesize natural speech from features extracted by the frozen Wav2Vec2.0 encoder, using a multi-receptive field fusion module and adversarial training with discriminators. This stage uses the LibriSpeech corpus to enhance speech representation learning. Stage 2: A lightweight LSTM adaptor maps neural activity to speech representations, enabling the frozen HiFi-GAN generator to re-synthesize high acoustic fidelity speech from neural data. The linguistic pathway involves a Transformer adaptor refining neural features to align with word tokens, which are fed into the frozen Parler-TTS model to generate high linguistic intelligibility speech. The voice cloning stage uses CosyVoice 2.0 (fine-tuned on TIMIT) to clone the speaker’s voice, ensuring the ultimate re-synthesized speech waveform matches the original stimuli in clarity and voice characteristics.

The acoustic pathway aims to synthesize speech with high acoustic naturalness, including prosody, speaker identity, and timbre, through a two-stage training process. In stage 1, we pre-train a high-fidelity generative adversarial network (HiFi-GAN) generator using a large corpus of natural speech data, without requiring paired neural recordings. This stage involves a speech autoencoder architecture consisting of a frozen Wav2Vec2.0 encoder (21), which transforms ECoG signals into latent representations known to correlate with brain activity (23, 24), and a trainable HiFi-GAN decoder, which synthesizes natural-sounding speech waveforms from these representations (Fig. 1B). HiFi-GAN (40), a generative adversarial network, was pre-trained to reconstruct high-fidelity speech using features extracted by the Wav2Vec2.0 encoder. This stage defines a rich, brain-aligned acoustic feature space (25) that does not require paired neural data. In stage 2, in order to decode speech from the brain, we next trained a lightweight acoustic adaptor that maps ECoG recordings onto the pre-defined latent space of the speech generator. This adaptor, a three-layer bidirectional long-short term memory (LSTM) model (∼9.4M parameters), was trained using only 20 minutes of neural recordings per participant while listening to natural speech. The HiFi-GAN waveform decoder remained frozen, allowing the adaptor to be trained efficiently and independently. This acoustic pathway enables high-fidelity speech reconstruction directly from neural signals, capturing fine-grained auditory detail while remaining data-efficient.

While the acoustic pathway ensures naturalness, it lacks explicit linguistic structure. The goal of the linguistic pathway is to decode discrete word-level representations from brain activity, facilitating the reconstruction of speech that is not only fluent but also semantically and syntactically accurate. To extract high-level linguistic information, we trained a Transformer-based linguistic adaptor (∼10.1M parameters) to align ECoG signals with word token representations derived from transcribed speech. This adaptor comprises a three-layer encoder and decoder and is trained in a supervised manner using the same 20-minute neural dataset. The adaptor output is fed into Parler-TTS (41), a state-of-the-art TTS model (∼880M parameters), which synthesizes fluent speech from word token sequences. This pathway captures syntactic structure and lexical identity that are not explicitly represented in the acoustic stream.

To generate speech that combines the strengths of both pathways, retaining the acoustic richness from the first and the intelligibility from the second, we fused the outputs in a final synthesis stage. The resulting waveforms were passed through CosyVoice 2.0 (42), a high-quality speech synthesis model with robust voice cloning and denoising capabilities. CosyVoice 2.0 was fine-tuned on the training set of TIMIT to improve articulation clarity, speaker-specific characteristics, and prosodic expressiveness. The final output preserves both low-level acoustic fidelity (e.g., pitch, prosody, and timbre) and high-level linguistic content (e.g., accurate word sequences), resolving the traditional trade-off in brain-to-speech decoding. Across the entire framework, unsupervised pretraining and supervised neural adaptation are used synergistically to achieve efficient, high-quality speech reconstruction.

The acoustic and linguistic performance of the reconstructed speech

We collected intracranial neural recordings from nine participants (4 male and 5 females; age range: 31-55) undergoing clinical neurosurgical evaluation for epilepsy. Each participant was implanted with high-density ECoG electrode arrays placed over the superior temporal gyrus and surrounding auditory speech cortex, based on clinical needs. During the experiment, participants passively listened to a set of 599 naturalistic English sentences spoken by multiple speakers. Neural signals were recorded at high temporal resolution while preserving the spatial specificity necessary to capture cortical responses to speech. This yielded approximately 20 minutes of labelled neural data for each participant. We then evaluated the quality of the re-synthesized sentences for each participant in the test set. Both subjective and objective evaluation metrics were adopted to give a comprehensive quantification of the effectiveness of this dual-path strategy in reconstructing intelligible and naturalistic speech from limited neural data. This included mean opinion score (MOS), mel-spectrogram correlation, word error rate (WER), and phoneme error rate (PER).

First, we evaluated the lower-level acoustic quality of the reconstructed speech. The average mel-spectrogram R2 (Fig. 2B) for the neural-driven re-synthesized speech across participants was 0.824 ± 0.029 (mean ± s.e., the best performance across participants achieving 0.844 ± 0.028, Fig. S3B). For comparison, we also computed these metrics on surrogate speech with additive noise (43) (measured in dB). The quality of the reconstructed speech was comparable to the R2 of −5dB and 0dB additive noise (−5dB: mel-spectrogram R2 = 0.864 ± 0.019, p = 0.161, two-sided t-test; 0dB: mel-spectrogram R2 = 0.771 ± 0.014, p = 0.064, two-sided t-test). The average MOS (Fig. 2c) for the neural-driven re-synthesized speech across participants was 3.956 ± 0.173, with the best performance across participants reaching 4.200 ± 0.119 (Fig. S3C).

Examples of neural-driven speech reconstruction performance.

(A) Ground truth and decoded speech comparisons. Waveforms (time vs. amplitude) and mel-spectrograms (time vs. frequency range 0-8 kHz) are shown for both the original (top) and reconstructed (bottom) speech samples, demonstrating preserved spectral-temporal patterns in the neural-decoded output. Decoding demonstration of phonemes and words is attached to the speech. (B) Mel-spectrogram correlation analysis. The KDE curve shows the distribution of aligned correlation coefficients between the original and reconstructed mel-spectrograms across all test samples, while the bar chart represents the percentage of trials in the test set (mean = 0.824 ± 0.028). Higher values reflect better acoustic feature preservation in the time-frequency domain. Purple dashed lines (light to dark) show average R² after adding −10 dB, −5 dB, and 0 dB additive noise to the original speech. (C) Human subject evaluations. The KDE curve displays the distribution of human evaluation results, while the bar chart represents the percentage of trials in the test set (mean = 3.956 ± 0.175). Higher values reflect better intelligibility. Purple dashed lines (light to dark) show the average MOS after adding −20 dB, −10 dB, and 0 dB additive noise to the original speech. (D-E) Word Error Rate (WER) and Phoneme Error Rate (PER) assessment.

Second, we further evaluated the linguistic content and intelligibility of the re-synthesized speech using word error rate (WER) and phoneme error rate (PER) metrics (Fig. 2D, E). The neural-driven re-synthesized speech achieved an average WER of 18.9 ± 3.3% across participants, with the best performance reaching 12.1 ± 2.1% (Fig. S3D). The WER result is comparable to adding −5dB noise on the original input speech stimuli (14.0 ± 2.1%, p = 0.332, two-sided t-test). Similarly, the average PER was 12.0 ± 2.5%, with the best performance reaching 8.3 ± 1.5% (Fig. S3E), comparable to adding −5dB noise on the original input speech stimuli (6.0 ± 1.5%, p = 0.068, two-sided t-test).

To evaluate the data efficiency of our framework and its sensitivity to the quantity of available neural recordings, we performed an ablation study by training models on progressively smaller subsets of the full per-subject training data (25%, 50%, 75%, and 100%). As shown in Supplementary Fig. S4, all three key performance metrics, acoustic fidelity (mel-spectrogram R², Fig. S4A), lexical accuracy (WER, Fig. S4B), and phonetic precision (PER, Fig. S4C), improved monotonically with increased training data volume. evaluation. The KDE curve displays distribution, while the bar chart represents the percentage of trials in the test set (mean WER = 0.189 ± 0.033, mean PER = 0.120 ± 0.025). Lower values indicate better word-level reconstruction accuracy. Purple dashed lines (light to dark) show average WER after adding −10 dB, −5 dB, and 0 dB additive noise to the original speech.

The acoustic pathway reconstructs high-fidelity lower-level acoustic information

The acoustic pathway, implemented through a bi-directional LSTM neural adaptor architecture (Fig. 1B), specializes in reconstructing fundamental acoustic properties of speech. This module directly processes neural recordings to generate precise time-frequency representations, focusing on preserving speaker-specific spectral characteristics like formant structures, harmonic patterns, and spectral envelope details. Quantitative evaluation confirms its core competency: achieving a mel-spectrogram R² of 0.793 ± 0.016 (Fig. 3B) demonstrates remarkable fidelity in reconstructing acoustic microstructure. This performance level is statistically indistinguishable from original speech degraded by 0dB additive noise (0.771 ± 0.014, p = 0.242, one-sided t-test). We chose a bidirectional LSTM architecture for this adaptor because its recurrent nature is particularly suited to modeling the fine-grained, short- to medium-range temporal dependencies (e.g., within and between phonemes and syllables) that are critical for acoustic fidelity. An ablation study comparing LSTM against Transformer-based adaptors for this task confirmed that LSTMs yielded superior mel-spectrogram reconstruction fidelity (higher R²), as detailed in Table S1, likely by avoiding the over-smoothing of spectrotemporal details sometimes induced by the strong global context modeling of Transformers.

Performance comparison among re-synthesized speech and MLP-regression, acoustic and linguistic baselines.

(A) Speech waveform and mel-spectrogram observations. Depicted are waveforms (time vs. amplitude) and mel-spectrograms (time vs. frequency ranging 0-8 kHz) for illustrative speech samples (Ground truth, MLP regression and acoustic-linguistic pathway intermediate outputs as baseline 1-3, and our ultimate re-synthesized natural speech). (B) Objective evaluation using mel-spectrogram R². Violin plots show the aggregated distribution of R² scores (0-1 scale) assessing spectral fidelity. The white dot represents the median, the box spans the interquartile range, and whiskers extend to ±1.5×IQR. Three dashed lines (light to dark) show average R² after adding −10 dB, −5 dB, and 0 dB additive noise to the original speech. (C) Subjective quality evaluation using mean opinion score (MOS). Violin plots show aggregated MOS distribution (1-5 scale). Three dashed lines (light to dark) indicate average MOS after adding −20 dB, −10 dB, and 0 dB additive noise. Note: The higher MOS for Baseline 3 (linguistic pathway) compared to ground truth is due to the superior acoustic quality of the modern speech corpus used to train Parler-TTS, whereas the TIMIT corpus contains inherent noise. GT: Ground truth. (D) Intelligibility assessment using word error rate (WER). Violin plots show aggregated WER distribution (0-1 scale). Three purple dashed lines (light to dark) show the average WERs after adding −10 dB, −5 dB, and 0 dB additive noise. (E) Phoneme error rate (PER) assessment. Format and noise conditions identical to panel D. Statistical significance markers: *:p<0.01, *:*p<0.001, n.s.: not significant.

To confirm that the acoustic pathway’s output is causally dependent on the neural signal rather than the generative prior of the HiFi-GAN, we performed a control analysis in which portions of the input ECoG recording were replaced with Gaussian noise. When either the first half, second half, or the entirety of the neural input was replaced by noise, the mel-spectrogram R² of the reconstructed speech dropped markedly, corresponding to the corrupted segment (Fig. S5). This demonstrates that the reconstruction is temporally locked to the specific neural input and that the model does not ‘hallucinate’ spectrotemporal structure from noise. These results validate that the acoustic pathway performs genuine, input-sensitive neural decoding.

However, exclusive reliance on acoustic reconstruction reveals fundamental limitations. Despite excellent spectral fidelity, the pathway produces critically impaired linguistic intelligibility. At the word level, intelligibility remains unacceptably low (WER = 74.6 ± 5.5%, Fig. 3D), while MOS and phoneme-level precision fares only marginally better (MOS = 2.878 ± 0.205, Fig. 3C; PER = 28.1 ± 2.2%, Fig. 3E). These deficiencies manifest perceptually as distorted articulation, inconsistent prosody, and compromised phonetic distinctiveness - collectively rendering the output functionally unintelligible for practical communication. This performance gap underscores a fundamental constraint: accurate acoustic feature reconstruction alone is insufficient for generating comprehensible speech, necessitating complementary linguistic processing.

The linguistic pathway reconstructs high-intelligibility, higher-level linguistic information

The linguistic pathway, instantiated through a pre-trained TTS generator (Fig. 1B), excels in reconstructing abstract linguistic representations. This module operates at the phonological and lexical levels, converting discrete word tokens into continuous speech signals while preserving prosodic contours, syllable boundaries, and phonetic sequences. It achieves a mean opinion score of 4.822 ± 0.086 (Fig. 3C) - significantly surpassing even the original human speech (4.234 ± 0.097, p = 6.674×10⁻33) in that the TIMIT corpus, recorded decades ago, contains inherent acoustic noise and relatively lower fidelity compared to modern speech corpus. Complementing this perceptual quality, objective intelligibility metrics confirm outstanding performance: WER reaches 17.7 ± 3.2%, with PER at 11.0 ± 2.3%. These results validate the pathway’s capacity for reconstructing high-level linguistic structures from symbolic representations. We employed a Transformer-based Seq2Seq architecture for this adaptor to effectively capture the long-range contextual dependencies across a sentence, which are essential for resolving lexical ambiguity and syntactic structure during word token decoding. This choice was validated by an ablation study (Table S2), indicating that the Transformer adaptor outperformed an LSTM-based counterpart in word prediction accuracy.

Nevertheless, this linguistic feature caption comes at an acoustic cost. The pathway exhibits significant limitations in spectral fidelity, achieving only 0.627 ± 0.026 (Fig. 3B) mel-spectrogram R² - substantially below the 0dB noise reference (0.771 ± 0.014, p = 2.919×10-8). This deficiency stems from fundamental architectural constraints: lacking explicit acoustic feature modelling, the system fails to preserve critical speaker-specific attributes like vocal tract resonances, glottal source characteristics, and individual articulatory patterns. Consequently, while linguistically intelligible, the output lacks authentic speaker identity and natural timbral qualities essential for realistic speech reconstruction.

The fine-tuned voice cloner further enhances the fidelity and intelligibility of re-synthesized speech

Voice cloning is used to bridge the gap between acoustic fidelity and linguistic intelligibility in speech reconstruction. This approach strategically combines the strengths of complementary pathways: the acoustic pathway preserves speaker-specific spectral characteristics while the linguistic pathway maintains lexical and phonetic precision. By integrating these components through neural voice cloning, we achieve balanced reconstruction that overcomes the limitations inherent in isolated systems. CosyVoice 2.0, the voice cloner module serves specifically as a voice cloning and fusion engine, requiring two inputs: (1) a voice reference speech (provided by the denoised output of the acoustic pathway) to specify the target speaker’s identity, and (2) a target word sequence (transcribed from the output of the linguistic pathway) to specify the linguistic content. The standalone baseline outputs of the two pathways can be integrated in this way.

The integration paradigm resolves this fundamental dichotomy through strategic combination. Where the acoustic pathway preserves spectral patterns (R² = 0.793 ± 0.016) but suffers unintelligibility (WER = 74.6 ± 5.5%), and the linguistic pathway delivers lexical precision but lacks acoustic fidelity, their fusion creates a complementary system. This synergistic architecture maintains the linguistic pathway’s exceptional intelligibility while incorporating the acoustic pathway’s spectrotemporal reconstruction. The resultant output achieves what neither can accomplish independently: linguistically precise speech with an authentic acoustic signature, effectively decoupling and optimizing both reconstruction dimensions within a unified framework.

To validate this integrated approach, we established three baseline models for comprehensive benchmarking. Baseline 1 (MLP-regression) directly maps neural recordings to mel-spectrograms without specialized processing. Baseline 2 represents the output of our acoustic pathway implementation, optimized for spectral feature reconstruction. Baseline 3 corresponds to the linguistic pathway output, generated by the TTS system without speaker-specific timbre adaptation.

The integration breakthrough occurs when combining this acoustic foundation with linguistic pathway components. The full pipeline achieves transformative improvements: WER decreases by 75% (18.9 ± 3.3% vs 74.6 ± 5.5%, p = 4.249×10-88, one-sided t-test, Fig. 2D), while PER drops by 58% (12.0 ± 2.5% vs 28.1 ± 2.2%, p = 2.592×10-44, one-sided t-test, Fig. 2E). Crucially, this intelligibility enhancement occurs without compromising acoustic fidelity: Mel spectrum R² = 0.824±0.029 vs 0.793 ± 0.016 (acoustic pathway, p = 4.486×10-3, one-sided t-test).

To further elucidate the phoneme recognition performance, we conducted an in-depth analysis of the errors present in the transcribed phoneme sequences derived from both original and re-synthesized speech. Specifically, we focused on the re-synthesized speech, where individual characters were subjected to insertion, deletion, or substitution operations to align them with the ground truth. This process not only quantified the errors but also provided insights into the relationships between different phonemes.

The confusion matrices (Fig. 4 A-D) illustrate the distribution of these errors. The diagonal elements represent the correct matches, indicating the accuracy for each phoneme, while off-diagonal non-zero elements denote substitution errors. Empty rows signify insertion errors (extraneous phonemes), and empty columns indicate deletion errors (missing phonemes). These visual representations highlight the specific challenges faced by our proposed model and the baseline models in accurately recognizing certain phonemes.

Phoneme recognition performance: confusion matrices and accuracy comparison

(A) Confusion matrix between transcribed phoneme sequences using our proposed model (horizontal axis) and ground truth (vertical axis). Diagonal values represent recognition accuracy for each phoneme (correct matches), while off-diagonal non-zero elements indicate substitution errors. Empty rows denote insertion errors (extraneous phonemes), empty columns indicate deletion errors (missing phonemes). All non-zero elements are highlighted for visual clarity. (B-D) Phoneme confusion matrix using baseline 1-3 for phoneme transcription. (E) Phoneme class clarity (PCC) across methods. PCC measures the proportion of mis-decoded phonemes that are confused within the same class (vowel-vowel or consonant-consonant) versus across classes (vowel-consonant). A higher PCC indicates that errors tend to be phonologically similar sounds, which supports intelligibility. The comparable PCC between our final integrated model and Baseline 3 (linguistic pathway) suggests that the phoneme-level error structure of our output is largely inherited from the high-quality linguistic prior embedded in the pre-trained TTS model (Parler-TTS). The violin plots show the distribution of PCC values across test samples. Asterisks (*) indicate phonemes where our proposed framework significantly outperforms baselines.

In addition to the confusion matrices, we categorized the phonemes into vowels and consonants to assess the phoneme class clarity. We defined “phoneme class clarity” (PCC) as the proportion of errors where a phoneme was misclassified within the same class versus being misclassified into a different class. The purpose of introducing PCC is to demonstrate that most of the misidentified phonemes belong to the same category (confusion between vowels or consonants), rather than directly comparing the absolute accuracy of phoneme recognition. For instance, a vowel being mistaken for another vowel would be considered a within-class error, whereas a vowel being mistaken for a consonant would be classified as a between-class error.

Our analysis revealed that the re-synthesized speech achieved a phoneme class clarity of 2.462 ± 0.201 (Fig. 4E), significantly outperforms chance level (1.360, calculated under the assumption that substituted phonemes are randomly distributed among all other phonemes, p = 3.424×10-6, one-sided t-test), MLP regression and acoustic pathway output (MLP regression: PCC = 1.694 ± 0.134, p = 1.312×10-3, one-sided t-test; acoustic pathway output: PCC = 1.735 ± 0.140, p = 2.367×10-3; linguistic pathway output: PCC = 2.446 ± 0.203, p = 0.477, one-sided t-test), indicating that most errors occurred within the same phoneme class. This result suggests that the model has a relatively good understanding of the phonetic structure, as it tends to confuse similar sounds rather than completely unrelated ones. Notably, this metric did not show significant differences compared to the baselines, suggesting that all models exhibit similar tendencies in terms of phoneme class confusion.

These comprehensive results confirm that our voice cloning framework successfully integrates the complementary strengths of acoustic and linguistic processing pathways. It achieves this integration without compromising either dimension: while significantly enhancing acoustic fidelity beyond what the linguistic pathway alone can achieve, it simultaneously maintains near-optimal linguistic intelligibility that the isolated acoustic pathway cannot provide. This balanced performance profile demonstrates the efficacy of our approach in producing high-quality, intelligible speaker-specific speech reconstruction.

Discussion

In this study, we present a dual-path framework that synergistically decodes both acoustic and linguistic speech representations from ECoG signals, followed by a fine-tuned zero-shot text-to-speech network to re-synthesize natural speech with unprecedented fidelity and intelligibility. Crucially, by integrating large pre-trained generative models into our acoustic reconstruction pipeline and applying voice cloning technology, our approach preserves acoustic richness while significantly enhancing linguistic intelligibility beyond conventional methods. Our dual-pathway architecture, while inspired by converging neuroscience insights on speech and language perception, was principally designed and validated as an engineering solution. The primary goal to build a practical decoder that achieves state-of-the-art reconstruction quality with minimal data. The framework’s success is therefore ultimately judged by its performance metrics, high intelligibility (WER, PER), acoustic fidelity (mel-spectrogram R²), and perceptual quality (MOS), which directly address the core engineering challenge we set out to solve. Using merely 20 minutes of ECoG recordings, our model achieved superior performance with a WER of 18.9% ± 3.3% and PER of 12.0% ± 2.5% (Fig. 2D, E). This integrated architecture, combining pre-trained acoustic (Wav2Vec2.0 and HiFi-GAN) and linguistic (Parler-TTS) models through lightweight neural adaptors, enables efficient mapping of ECoG signals to dual latent spaces. Such methodology substantially reduces the need for extensive neural training data while achieving breakthrough word clarity under severe data constraints. The results demonstrate the feasibility of transferring the knowledge embedded in speech-data pre-trained artificial intelligence (AI) models into neural signal decoding, paving the way for more advanced brain-computer interfaces and neuroprosthetics.

Our framework establishes a framework for speech decoding by outperforming prior acoustic-only or linguistic-only approaches (Table S3) through integrated pretraining-powered acoustic and linguistic decoding. This dual-path methodology proves particularly effective where traditional methods fail, as it simultaneously resolves the acoustic-linguistic trade-off that has long constrained reconstruction quality. While end-to-end re-synthesis remains intuitively appealing, prior work confirms that direct methods achieve only modest performance given neural data scarcity(15). To overcome this, we propose a hybrid encoder-decoder architecture utilizing: (1) pre-trained spectral synthesizers (Wav2Vec2.0 and HiFi-GAN) for acoustic fidelity, and (2) transformer-based token decoders (Parler-TTS) for linguistic precision. Participant-specific projection modules further ensure cross-subject transferability with minimal per-user calibration. Collectively, these advances surmount core limitations of direct decoding, enabling unprecedented speech quality within extreme data constraints.

A pivotal advancement lies in establishing robust, clinically relevant intelligibility metrics, notably achieving 18.9% WER and 12.0% PER through standardized evaluation, directly addressing the core challenge of word recognition in speech decoding. Our dual-path framework enables comprehensive sentence-level assessment through objective benchmarks (phonetic precision via PER and lexical accuracy via WER evaluated by Whisper ASR), acoustic fidelity validation (high mel-spectrogram correlation: 0.824 ± 0.029, Fig. 2B) and human perceptual testing (near-‘excellent’ MOS ratings: 3.956 ± 0.173, Fig. 2C). Critically, this tripartite evaluation spanning acoustic (spectral/time-domain), linguistic (phoneme/word), and perceptual dimensions revealed superior reconstruction quality, while objective metrics confirmed spectral coherence rivalling clean speech inputs.

The phoneme confusion pattern observed in our model output (Fig. 4A) differs from classic human auditory confusion matrices. We attribute this divergence primarily to the influence of the Parler-TTS model, which serves as a strong linguistic prior in our pipeline. This component is trained to generate canonical speech from text tokens. When the upstream neural decoding produces an ambiguous or erroneous token sequence, the TTS model’s internal language model likely performs an implicit ‘error correction,’ favoring linguistically probable words and pronunciations. This underscores that our model’s errors arise from a complex interaction between neural decoding fidelity and the generative biases of the synthesis stage.

Our findings demonstrate that neural representations exhibit dual alignment, which is not only with acoustic features from deep speech models (Wav2Vec2.0), but critically with linguistic features from language models (Parler-TTS), establishing a bidirectional bridge between cortical activity and hierarchical speech representations. This convergence mirrors the complementary processing streams observed in the human speech cortex (24, 28) where self-supervised models capture both spectrotemporal patterns and semantic structures. Such unified alignment marks a paradigm shift in brain-AI integration: The discovery of near-linear mappings between neural and multimodal AI spaces unlocks transformative applications in speech synthesis and cognitive interfaces. Furthermore, our results confirm that foundation models serve as scalable ‘cognitive mirrors’ (44). With the advent of more sophisticated generative models, we anticipate enhanced neural decoding capabilities, including the potential to improve signal quality (in terms of SNR) and refine the generative model itself. Additionally, this framework extends beyond speech to other modalities, such as vision, suggesting that similar principles may apply to the generation of visual content from neural signals.

This work advances speech decoding and potentially speech BCIs by enabling real-time speech reconstruction with minimal neural data, improving scalability and practicality through pre-trained AI models, and suggesting potential for broader applications in perception, memory, and emotion. By adapting neural activity to a common latent space of the pre-trained speech autoencoder, our framework significantly reduces the quantity of neural data required for effective speech decoding, thereby addressing a major limitation in current BCI technologies. This reduction in data needs paves the way for more accessible and scalable BCI solutions, particularly for individuals with speech impairments who stand to benefit from immediate and intelligible speech reconstruction. Furthermore, the applicability of our model extends beyond speech, hinting at the possibility of decoding other cognitive functions once the corresponding neural correlates are identified. This opens up exciting avenues for expanding BCI functionality into areas such as perception, memory, and emotional expression, thereby enhancing the overall quality of life for users of neuroprosthetic devices.

There are several limitations in our study. The quality of the re-synthesized speech heavily relies on the performance of the generative model, indicating that future work should focus on refining and enhancing these models. Currently, our study utilized English speech sentences as input stimuli, and the performance of the system in other languages remains to be evaluated. Regarding signal modality and experimental methods, the clinical setting restricts us to collecting data during brief periods of awake neurosurgeries, which limits the amount of usable neural activity recordings. Overcoming this time constraint could facilitate the acquisition of larger datasets, thereby contributing to the re-synthesis of higher-quality natural speech. Furthermore, the inference speed of the current pipeline presents a challenge for real-time applications. On our hardware (a single NVIDIA GeForce RTX 3090 GPU), synthesizing speech from neural data takes approximately two to three times longer than the duration of the target speech segment itself. This latency is primarily attributed to the sequential processing in the autoregressive linguistic adaptor and the computationally intensive high-fidelity waveform generation in the vocoder (CosyVoice 2.0). While the current study focuses on offline reconstruction accuracy, achieving real-time or faster-than-real-time inference is a critical engineering goal for viable speech BCI prosthetics. Future work must therefore prioritize architectural optimizations, such as exploring non-autoregressive decoding strategies and more efficient neural vocoders, alongside potential hardware acceleration. Additionally, exploring non-invasive methods represents another frontier; with the accumulation of more data and the development of more powerful generative models, it may become feasible to achieve effective non-invasive neural decoding for speech re-synthesis. Moreover, while our framework adopts specialized architectures (LSTM and Transformer) for distinct decoding tasks, an alternative approach is to employ a unified multimodal large language model (LLM) capable of joint acoustic-linguistic processing. Finally, the current framework requires training participant-specific adaptors, which limits its immediate applicability for new users. A critical next step is to develop methods that learn a shared, cross-participant neural feature encoder, for instance, by applying contrastive or self-supervised learning techniques to larger aggregated ECoG datasets. Such an encoder could extract subject-invariant neural representations of speech, serving as a robust initialization before lightweight, personalized fine-tuning. This approach would dramatically reduce the amount of per-subject calibration data and time required, enhancing the practicality and scalability of the decoding framework for real-world BCI applications.

In summary, our dual-path framework achieves high speech reconstruction quality by strategically integrating language models for lexical precision and voice cloning for vocal identity preservation, yielding a 37.4% improvement in MOS scores over conventional methods. This approach enables high-fidelity, sentence-level speech synthesis directly from cortical recordings while maintaining speaker-specific vocal characteristics. Despite current constraints in generative model dependency and intraoperative data collection, our work establishes a new foundation for neural decoding development. Future efforts should prioritize: (1) refining few-shot adaptation techniques, (2) developing non-invasive implementations, (3) expanding to dynamic dialogue contexts, and (4) cross-subject applications. The convergence of neurophysiological data with multimodal foundation models promises transformative advances, not only revolutionizing speech BCIs but potentially extending to cognitive prosthetics for memory augmentation and emotional communication. Ultimately, this paradigm will deepen our understanding of neural speech processing while creating clinically viable communication solutions for those with severe speech impairments.

Materials and Methods

Participants

The study comprised 9 monolingual right-handed participants, each with electrodes placed on the left hemisphere cortex to clinically monitor seizure activities. Electrode placement followed specific clinical requirements (see Fig. S1 for grid placement in each placement). Participants were thoroughly informed about the experiment, as outlined in the consent document approved by the Institutional Review Board (IRB) of the University of California, San Francisco. They volunteered for participation, ensuring that their involvement did not influence their clinical care. Additionally, verbal consent was obtained from participants at the beginning of each experimental session.

Data acquisition and neural signal processing

All participants utilized high-density ECoG grids of uniform specifications and type. During the experimental tasks, neural signals captured by the ECoG grids were acquired using a multi-channel amplifier optically linked to the digital signal processor. Data recording was performed through Tucker-Davis Technology (TDT) OpenEx software. The local field potential at each electrode contact was amplified and sampled at 3052 Hz. Subsequent to data collection, the experiment employed the Hilbert transform to compute the analytic amplitudes of eight Gaussian filters (with center frequencies ranging from 70 to 150 Hz), and the signal was down-sampled to 50 Hz. These tasks were segmented into record blocks lasting approximately 5 minutes each. The neural signals were z-score normalized for each recording block.

Experimental Stimuli

The acoustic stimuli employed in this study comprised natural and continuous English speech. The English speech stimuli were sourced from the TIMIT corpus (45), consisting of 499 sentences read by 286 male and 116 female speakers. A silence interval of 0.4 seconds was maintained between sentences. The task was organized into five blocks, each with a duration of approximately 5 minutes.

Electrode localization

In cases involving chronic monitoring, the electrode placement procedure involved pre-implantation MRI and post-implantation CT scans. In awake cases, temporary high-density electrode grids were utilized to capture cortical local potentials, with their positions recorded using the Medtronic neuronavigation system. The recorded positions were then aligned to pre-surgery MRI data, with intraoperative photographs serving as additional references. Localization of the remaining electrodes was achieved through interpolation and extrapolation techniques.

Speech-responsive electrode selection

The selection of speech-responsive electrodes played a crucial role in the re-synthesis of high-quality natural speech and in avoiding overfitting. Onsets were identified as the initiation of speech and are preceded by more than 400 ms of silence. A paired sample t-test was conducted to investigate whether the average post-onset response (400 ms to 600 ms) exhibited a significant increase compared to the average pre-onset response (−200 ms to 0 ms, p < 0.01, one-sided, Bonferroni corrected).

Architecture and training of the speech Autoencoder

The speech Autoencoder served as a pivotal component, taking natural speech as input and reproducing the input itself. Within this model, the intermediate encoding layers were construed as the feature extraction layers. The architecture comprises an encoder, namely Wav2Vec2.0, and a decoder, which is the HiFi-GAN generator.

The Wav2Vec2.0 encoder was comprised of 7 1D convolutional layers that down-sample 16kHz natural speech to 50Hz, thereby extracting 768-dimensional local features. Additionally, it integrated 12 transformer-encoder blocks to extract contextual feature representations from unlabeled speech data.

On the other hand, the HiFi-GAN generator incorporated 7 transposed convolutional layers for up-sampling, along with multi-receptive field (MRF) fusion modules. Each MRF module aggregated the output features from multiple ResBlocks. The discriminator within the HiFi-GAN encompassed both multi-scale and multi-period discriminators. Notably, the up-sample rates and kernel sizes of each ResBlock in the MRFs were meticulously configured. The up-sample rates of each ResBlock in the MRFs were set as 2, 2, 2, 2, 2, 2, 5 respectively. The up-sample kernel sizes of each ResBlock in the MRFs were set as 2, 2, 3, 3, 3, 3, 10. The periods of the multi-period discriminator were set as 2, 3, 5, 7, 11.

The HiFi-GAN generator 𝐺 and discriminators 𝐷k were trained simultaneously and adversarially. The loss function consisted of three parts: adversarial loss 𝐿Adv to train the HiFi-GAN, mel-spectrogram loss 𝐿Mel’to improve the training efficiency of the generator and feature matching loss 𝐿FM to measure the similarity between original speech and re-synthesized speech from learned features: The regularization coefficients 𝜆FM and 𝜆Mel’were set as 2 and 50 respectively in the implementation. In detail, the three terms of 𝐿) were: Where 𝜑 represented the mel-spectrogram operator, 𝑇 was the number of layers in the discriminator, 𝐷i and 𝑁i were the features and the number of features in the 𝑖th layer of the discriminator, 𝑥 and 𝑥SAE represented the original speech and the reconstructed speech using the pre-trained speech Autoencoder: The parameters in Wav2Vec2.0 were frozen within this training phase. The parameters in HiFi-GAN were optimized using the Adam optimizer with a fixed learning rate of 10-5, 𝛽1 = 0.9, 𝛽2 = 0.999. We trained this Autoencoder in LibriSpeech, a 960-hour English speech corpus with a sampling rate of 16kHz, which is entirely separate from the TIMIT corpus used for our ECoG experiments. We spent 12 days in parallel training on 6 Nvidia GeForce RTX3090 GPUs. The maximum training epoch was 2000. The optimization did not stop until the validation loss no longer decreased.

Acoustic feature adaptor training and ablation test

The acoustic feature adaptor was trained to acquire the acoustically fidelity re-synthesized natural speech. The acoustic speech re-synthesizer received 𝑧-scored high gamma ECoG snippets with dimensions 𝑁 × 𝑇, where 𝑁 was the number of selected speech-responsive electrodes, and 𝑇 was the ECoG recordings (50 Hz) down-sampled to the same sample rate as the speech features extracted from pre-trained Wav2Vec2.0.

A three-layer bidirectional LSTM adaptor 𝑓 (see Fig. 1B) aligned recorded ECoG signals to speech features. The output of the adaptor was 768 × 𝑇, as the input of the HiFi-GAN generator pre-trained in phase 1. The ultimate output of the HiFi-GAN was a speech waveform with a length of 1 × (320𝑇 + 80) in a 16K sample rate.

We froze all the parameters of the pre-trained speech Autoencoder while training the adaptor. The loss function used in this phase consisted of two parts: the mel-spectrogram loss 𝐿Mel and the Laplacian loss 𝐿Lap: where the regularization coefficient 𝜆 was set as 0.1 using the ablation test.

𝐿Mel was used to enhance the acoustic fidelity and intelligibility of the ultimate re-synthesized speech from recorded neural activity: where 𝑥 and 𝑥a respectively represented the original and acoustic fidelity speech, and 𝛷 represented the mel-spectrogram operator.

In order to improve the phonetic intelligibility, a Laplace operator as a convolution kernel was applied to the mel-spectrogram, which was a simplified representation of the convolution on the spectrogram, while convolution on the spectrogram was proven to be effective on speech denoising and enhancement(46): The parameters in the HiFi-GAN generator were frozen within this training phase. The parameters in the adaptor were trained in Stochastic Gradient Descent (SGD) optimizer with a fixed learning rate of 3 × 10-3 and a momentum of 0.5. A 10% dropout layer was added in this stage to avoid overfitting. The optimization did not stop until the validation loss no longer decreased.

For each participant, we trained a personalized feature adaptor. Each individual feature adapter underwent training for 500 epochs on one NVidia GeForce RTX3090 GPU, which required 12 hours to complete. The dataset was divided into training, validation, and test sets, comprising 70%, 20%, and 10% of the total trials, respectively. The choice of the LSTM architecture over alternatives was informed by an ablation study presented in Table S1.

Linguistic feature adaptor training and ablation test

The linguistic feature adaptor was designed to decode word token sequences from ECoG recordings. The model received z-scored high gamma ECoG snippets with dimensions (𝑁 × 𝑇), where 𝑁 was the number of selected speech-responsive electrodes, and 𝑇 was the ECoG recordings (50 Hz) down-sampled to the same sample rate as the acoustic feature training stage.

An attention-based Seq2Seq Transformer architecture (see Fig. 1B) was employed to adapt ECoG signals to linguistic features. The linguistic adaptor consisted of 3 encoder layers and 3 decoder layers, with a hidden dimension of 256 and 8 attention heads. The input ECoG signals were first projected into a latent space using a linear layer, followed by positional encoding to capture temporal dependencies. The decoder generated word token sequences autoregressively, starting with a start-of-sequence (SOS) token. The output dimension matched the linguistic feature space (1024-dimensional).

The loss function combined the token-level KL-divergence and sequence length prediction loss with L2 regularization: where 𝐿token measured the KL-divergence between predicted and target token distributions: and 𝐿length was a Huber loss between predicted and actual sequence lengths: The regularization coefficient 𝜆1 was set to 1 based on ablation tests, and L2 regularization (weight decay = 0.01) was applied to mitigate overfitting. The model was trained using the Adam optimizer with a fixed learning rate of 10-4 for 500 epochs. The dataset was split into training (70%), validation (20%), and test (10%) sets, consistent with the acoustic feature adaptor. For each participant, a personalized linguistic adaptor was trained on one NVidia GeForce RTX3090 GPU, requiring approximately 6 hours to complete. The Transformer architecture was selected based on its superior performance in word token prediction compared to recurrent networks, as detailed in Table S2.

Automatic phoneme recognition of re-synthesized speech sentences

In order to quantitatively evaluate the synthesized speech is to recognize the phonemes, we fine-tuned a phoneme recognizer using Wav2Vec2.0, a logistic regression phoneme classifier, and a phoneme decoder. Wav2Vec2.0 was used for feature extraction, the classifier outputs a probability vector from Wav2Vec2.0 features, and the phoneme decoder decoded a phoneme sequence from the probability vectors to finally get the phoneme sequence.

The models are trained and evaluated on the training set and test set of TIMIT, and tested on the re-synthesized ECoG recordings. These parameters were trained using the AdamW optimizer with a fixed learning rate of 10-4, 𝛽1= 0.9, 𝛽2 = 0.999, and 𝜀 = 10-8 in the CTC loss function. The optimization did not stop until the validation loss no longer decreased.

Phoneme error rate (PER) and word error rate (WER)

We implemented a two-stage transcription pipeline to evaluate linguistic intelligibility. First, word sequences were transcribed using OpenAI’s Whisper model(47) (base version) with default parameters, which provided both word-level alignments and confidence scores. The reference and decoded word sequences were then aligned using edit distance to compute WER. The error rates were calculated by applying the same edit distance algorithm to the forced aligned phoneme sequences. This combined approach ensured consistent phoneme representations between natural and synthesized speech during comparison: Where Edit(𝑜𝑟𝑖𝑔𝑖𝑛𝑎𝑙, 𝑡𝑟𝑎𝑛𝑠𝑐𝑟𝑖𝑏𝑒𝑑) denotes the edit distance between original and transcribed phoneme sequences; 𝑁 is the number of phonemes in the sentence.

Alignment of the ultimate re-synthesized speech to calculate the mel-spectrogram R2

To quantitatively evaluate the acoustic fidelity of the re-synthesized speech using mel-spectrogram R2, we performed temporal alignment between the reconstructed and ground truth mel-spectrograms. First, word-level timestamps were extracted from the transcribed speech using OpenAI’s Whisper model, which provided precise boundaries for voiced segments. The mel-spectrogram of the re-synthesized speech was then segmented into these intervals, excluding silent or non-speech regions. Each voiced segment was resampled to match the temporal resolution of the corresponding ground truth spectrogram using linear interpolation, ensuring frame-by-frame comparability. The squared Pearson correlation coefficient (R²) was computed over the aligned spectrograms, measuring the proportion of variance in the ground truth spectrogram explained by the reconstructed features. This approach accounted for natural variations in speech rate while rigorously quantifying spectral fidelity, with higher R² values (closer to 1) indicating superior preservation of time-frequency structures critical for speech intelligibility and naturalness.

Voice cloning for the ultimate re-synthesized natural speech

To achieve personalized speech output, we implemented a voice cloning pipeline that preserves the subject’s unique vocal features (Fig. S2). The system takes two key inputs: (1) a clean voice sample from the denoised linguistic speech reference, and (2) the corresponding word sequence transcribed by Whisper. This approach serves dual purposes - Whisper first transcribes the original speech to provide accurate text prompts, then later evaluates the intelligibility of the ultimate re-synthesized speech by transcribing it again for word error rate calculation.

The cloning process begins with extracting vocal fingerprints from the short reference sample, capturing distinctive timbral qualities like pitch contour and formant structure. These vocal features are then mapped to the target word sequences, generating synthetic speech that maintains the subject’s natural voice patterns while clearly articulating the desired content.

We specifically employed the “3-second rapid cloning” mode to optimize for both efficiency and voice quality. This ultra-fast processing allows for quick adaptation to individual speakers while still producing highly natural-sounding results. The synthesized output was rigorously evaluated through both objective metrics (Whisper-based transcription accuracy) and subjective listening tests to ensure faithful reproduction of the original voice features.

Fine-tuning CosyVoice2.0 on TIMIT corpus

To enhance the fidelity and intelligibility of re-synthesized speech while utilizing the generative capacity of large speech models for artifact mitigation, CosyVoice2.0 was fine-tuned on the sentences from the TIMIT corpus used in the training set of neural-driven dual-pathway model training. The architecture comprises two core components: (i) a “Flow” module converting tokens to acoustic features via Text Embedding, Speaker Embedding (linear layer), Encoder, and causal conditional CFM Decoder; and (ii) a “Hift” vocoder synthesizing waveforms from Flow outputs using an F0 extractor, Up-sampler, and 9 residual blocks. Selective parameter optimization was applied to: (i) the first and final layers of the 12-layer Flow decoder, and (ii) the Speaker Embedding linear layer. The model mapped speech autoencoder outputs and punctuation-free plain text to ground-truth waveforms, optimized with a compound loss consists of mel-spectrogram KL-divergence and MFCC loss: These models were trained using AdamW optimizer with a fixed learning rate of 10-5, and set 𝛽1 = 0.9, 𝛽2 = 0.999, and 𝜀 = 10-8. 𝑥 represented the speech waveform, and 𝛷 represented the mel-spectrogram operator.

Ablation study on training data volume

To assess the impact of training data quantity on decoding performance, we conducted an additional ablation experiment. For each participant, we created subsets of the full training set corresponding to 25%, 50%, and 75% of the original data by random sampling while preserving the temporal continuity of speech segments. Personalized acoustic and linguistic adaptors were then independently trained from scratch on each subset, following the identical architecture and optimization procedures described above. All other components of the pipeline, including the frozen pre-trained generators (HiFi-GAN, Parler-TTS) and the CosyVoice 2.0 voice cloner, remained unchanged. Performance metrics (mel-spectrogram R², WER, PER) were evaluated on the same held-out test set for all data conditions. The results (Fig. S4) demonstrate that when more than 50% of the training data is utilized, performance degrades gracefully rather than catastrophically, which is a promising indicator for clinical applications with limited data collection time.